Heterozygosity at a conserved candidate sex determination locus is associated with female development in the clonal raider ant (Ooceraea biroi)

Version 1

Abstract

Abstract

Sex determination is a developmental switch that triggers sex-specific developmental programs. This switch is “flipped” by the expression of genes that promote male- or female-specific development. Many lineages have evolved sex chromosomes that act as primary signals for sex determination. However, haplodiploidy (males are haploid and females are diploid), which occurs in ca. 12% of animal species, is incompatible with sex chromosomes. Haplodiploid taxa must, therefore, rely on other strategies for sex determination. One mechanism, “complementary sex determination” (CSD), uses heterozygosity as a proxy for diploidy. In CSD, heterozygosity at a sex determination locus triggers female development, while hemizygosity or homozygosity permits male development. CSD loci have been mapped in honeybees and two ant species, but we know little about their evolutionary history. Here, we investigate sex determination in the clonal raider ant, Ooceraea biroi. We identified a 46kb candidate CSD locus at which all females are heterozygous, but most diploid males are homozygous for either allele. As expected for CSD loci, the candidate locus has more alleles than most other loci, resulting in a peak of nucleotide diversity. This peak negligibly affects the amino acid sequences of protein-coding genes, suggesting that heterozygosity of a non-coding genomic sequence triggers female development. This locus is distinct from the CSD locus in honeybees but homologous to a CSD locus mapped in two distantly related ant species, implying that this molecular mechanism has been conserved since a common ancestor that lived approximately 112 million years ago.

Introduction

Whether an animal develops as a male or a female is typically determined by a switch early in development. This switch, known as sex determination, is a multistep process beginning with primary sex determination signals triggering sex-specific splicing of downstream transcription factors that encode sexual identity (Williams and Carroll 2009). Although the downstream transcription factors are evolutionarily conserved, primary sex determination signals are evolutionarily labile and comprise a diverse range of environmental and genetic mechanisms (Bachtrog et al. 2014). Genetically encoded primary signals (sex-determining genes or chromosomes) are usually found in only one sex. Such systems do not work in haplodiploid taxa (an estimated 12% of animal species, including ants (Normark 2003)), where any allele on any homologous chromosome can be transmitted from diploid females to their haploid male sons, and all alleles experience selection to be viably transmitted from both sexes (Whiting 1935). Haplodiploids, therefore, require alternative sex determination mechanisms, but we know little about the molecular details.

Some parasitoid wasps use a mechanism involving maternal imprinting (Verhulst et al. 2010; Zou et al. 2020). However, in many Hymenoptera, inbred crosses produce diploid males, which led to the hypothesis that female development is triggered by heterozygosity as a proxy for diploidy (Whiting 1933; Whiting 1943). Under this “complementary sex determination” (CSD) hypothesis, different alleles at a given sex determination locus “complement” one another, meaning that heterozygosity triggers female development (Fig. 1A). By contrast, the presence of a single allele (either because the individual is haploid or because a diploid individual is homozygous at the sex determination locus) permits male development. Under CSD, diploid males produce diploid sperm and, therefore, are functionally sterile (van Wilgenburg et al. 2006). This has several evolutionary consequences (Cook and Crozier 1995); at the population level, there is selection against homozygosity, often meaning obligatory avoidance of inbreeding (Page 1980). At the gene level, there is negative frequency-dependent (balancing) selection on CSD loci, resulting in the co-occurrence of many alleles at similar frequencies within populations (Laidlaw et al. 1956).

Theoretical expectations for complementary sex determination (CSD).

(A) Cartoon depicting a mating between a diploid female with two different alleles at a sex determination locus and a haploid male bearing one of these two alleles. Half of the sexually produced diploid offspring are expected to develop as males. (B) Table illustrating multi-locus CSD in diploids, in which heterozygosity of at least one sex determination locus is required for female development. By contrast, only homozygosity at all loci results in diploid male development. (C) Cartoon depicting how CSD might work in asexual species such as the clonal raider ant. If diploid males arise from losses of heterozygosity at sex loci, then homozygosity for either of the two alleles should trigger male development. Offspring proportions reflect empirical observations in the clonal raider ant (Kronauer et al. 2012; Oxley et al. 2014).

Having been a favored hypothesis for more than half a century, CSD was demonstrated in the honeybee Apis mellifera in 2003 (Beye et al. 2003). In A. mellifera, biallelic heteromers of the protein encoded by the complementary sex determiner (csd) gene trigger female development, whereas monoallelic homomers permit male development (Beye et al. 2003; Beye et al. 2013; Otte et al. 2023; Seiler and Beye 2024). Although this primary sex determination signal differs from primary signals in other insects, it nonetheless converges upon the pathway that transduces sex determination signals in holometabolous insects (Bopp et al. 2014; Sawanth et al. 2016; Wexler et al. 2019). The Csd protein sex-specifically splices the closely linked splicing factor feminizer (fem) (the honeybee version of transformer (tra) in Drosophila and other insects), which in turn leads to the production of sex-specific isoforms of doublesex (dsx), a transcription factor that regulates sex-specific development (Hasselmann et al. 2008; Gempe et al. 2009; Verhulst et al. 2010; Zou et al. 2020). Intriguingly, csd is a paralog of fem. Following the prediction that CSD loci should evolve under balancing selection, many csd alleles are held at similar frequencies, and the locus bears signatures of adaptive evolution (Hasselmann and Beye 2004), with amino acid-level heterozygosity in certain protein domains required for female development (Otte et al. 2023).

CSD is assumed to occur in many Hymenoptera that occasionally produce diploid males (van Wilgenburg et al. 2006). Until recently, however, heterozygosity-dependent female development had not been demonstrated outside of A. mellifera, which remained the only species for which a CSD locus had been genetically mapped. A recent genetic mapping study in the wasp Lysiphlebus fabarum identified at least one and as many as four candidate sex determination loci (Matthey-Doret et al. 2019) (the multi-locus extension of the CSD hypothesis proposes that heterozygosity at any of multiple sex determination loci is sufficient to trigger female development (Crozier 1971) (Fig. 1B)). However, a requirement of heterozygosity at these loci for female development was not demonstrated. In the ant Vollenhovia emeryi, a mapping study found two sex determination QTL (Miyakawa and Mikheyev 2015), with follow-up work suggesting that these function as a multi-locus CSD system (Miyakawa and Miyakawa 2023). One of these QTL (V.emeryiCsdQTL1) contains two closely linked tra homologs, reminiscent of the locus containing csd and fem in A. mellifera. By contrast, the other QTL (V.emeryiCsdQTL2) only contains genes without annotated functions in sexual development. Although V. emeryi sex determination is likely integrated through tra and dsx, the molecular basis of their primary sex determination signals remains unclear (Miyakawa et al. 2018; Miyakawa and Miyakawa 2023).

Recently, a sex determination locus was identified in the Argentine ant, Linepithema humile (Pan et al. 2024). Consistent with CSD, all pairwise heterozygous combinations of the seven identified alleles were found in females, whereas diploid males were invariably homozygous for one of the seven alleles. Moreover, these alleles occurred at roughly equal frequencies within populations, suggesting evolution under balancing selection. The mapped locus is located within a noncoding genomic region tightly linked to ant noncoding transformer splicing regulator (ANTSR), a long noncoding RNA (lncRNA) with previously uncharacterized function. ANTSR knockdown in female-destined embryos leads to male-specific splicing of tra, suggesting that this lncRNA transduces the primary sex determination signal encoded by the heterozygosity (or hemi- or homozygosity) of the linked noncoding region (Pan et al. 2024). Because the synteny of this region (including the presence of a lncRNA putatively homologous to ANTSR) is conserved across the ants, bees, and vespoid wasps, it was suggested that this sex determination locus may be deeply conserved among the Aculeata (Pan et al. 2024). However, synteny is insufficient to demonstrate functional conservation, and genetic mapping of sex determination has yet to be performed in most of these taxa.

To explore the evolution of sex determination across ants, we investigated the topic in the clonal raider ant, Ooceraea biroi. Diploid males occur sporadically in this species, suggesting that, like other ants, it might employ CSD (Kronauer et al. 2012). O. biroi reproduces via a mode of parthenogenesis in which two haploid nuclei from a single meiosis fuse to produce diploid offspring (central fusion automixis) (Oxley et al. 2014). Although heterozygosity is maintained with high fidelity despite meiotic crossover recombination (Lacy et al. 2024), losses of heterozygosity occur occasionally, probably due to the inheritance of one recombined and one non-recombined version of a homologous chromosome (Kronauer et al. 2012; Oxley et al. 2014; Trible et al. 2023; Lacy et al. 2024). Therefore, if O. biroi uses CSD, diploid males might result from losses of heterozygosity at sex determination loci (Fig. 1C). Here, we use whole genome sequencing to map a sex determination locus, demonstrate that most diploid males carry a loss of heterozygosity at this locus, assess the homology of this locus with other sex determination loci, and explore whether balancing selection has shaped evolution at this locus.

Results

Identification of a candidate sex determination locus on chromosome 4

To map sex determination loci in O. biroi, we first identified males based on sexual dimorphism and distinguished diploid from haploid males based on heterozygosity at eight genetic markers (Supplementary Table S1, File S1, and Fig. S1). We then sequenced the genomes of 16 diploid males and compared them with 19 previously sequenced genomes of diploid females (Trible et al. 2023; Lacy et al. 2024) (descriptions and metadata for all genomes sequenced in this study are in Supplementary Table S2). CSD loci must be heterozygous to trigger female development but can be homozygous in diploid males. To identify such sites, we calculated a “CSD index” for each single nucleotide polymorphism (SNP), which equals zero if any female is homozygous and equals the proportion of diploid males that are homozygous if all females are heterozygous. We found a peak of the CSD index on chromosome 4, which deviated significantly from the null distribution of CSD index values (Fig. 2A). At this peak, homozygosity levels differed significantly between males and females (supplementary fig. S2).

Whole genome sequencing reveals a candidate sex determination locus on chromosome 4.

(A) Karyoplot depicting the mean CSD index p-value in 50kb windows with a 15kb sliding interval. The CSD index peak is shown as a green dotted line. The significance threshold (p=0.05) and FDR-corrected significance threshold (p=0.002) are indicated by gold and brown lines, respectively. The grey histogram shows the number of ancestrally heterozygous SNPs in 300kb windows. (B) Stacked bar plots for all diploids used for mapping, with one horizontal line for each ordered putatively ancestrally heterozygous SNP on chromosome 4. For each sample, SNPs that retain ancestral heterozygosity are drawn as grey lines, whereas SNPs that have lost ancestral heterozygosity are drawn as black lines.

To corroborate our results, we looked for evidence of segmental losses of heterozygosity along chromosome 4. Eleven out of 16 sequenced diploid males bore segmental losses of heterozygosity on chromosome 4 (Fig. 2B). The intersection of these LOH segments comprised a 46kb peak at which all females were heterozygous, and most (11 out of 16) diploid males were homozygous. Under the CSD hypothesis, homozygosity for any allele at CSD loci permits male development. Consistently, O. biroi diploid males were homozygous for either allele at the 46kb region on chromosome 4 (Table 1, Supplementary Fig. S3).

The stretches of homozygosity found in diploid males could have resulted from segmental deletions resulting from improperly repaired DNA damage or rare copy-neutral losses of heterozygosity resulting from thelytokous parthenogenesis (Kronauer et al. 2012; Oxley et al. 2014; Trible et al. 2023; Lacy et al. 2024). To distinguish between these two scenarios, we analyzed read depth and found that runs of homozygosity were not accompanied by changes in copy number (Supplementary Fig. S3). This implies that diploid males arise when rare losses of heterozygosity, which result from crossover recombination during meiosis, span the CSD locus.

Surprisingly, five of the 16 diploid males were not homozygous in this 46kb region. One possible explanation is that these individuals had mutations in other genes involved in the sex-determination pathway. We therefore inspected all unique mutations and losses of heterozygosity in these five diploid males, but did not find any in genes with annotated sex determining functions (Supplementary File S2). These individuals may carry mutations or losses of heterozygosity in genes or regulatory elements with unannotated sex determination function. Alternatively, male development may have been triggered by rare stochastic perturbations to gene expression or splicing. We address these possibilities further in the Discussion.

The O. biroi CSD locus is homologous to another ant sex determination locus but not to honeybee csd

To assess homology to CSD loci mapped in other species, we identified O. biroi orthologs of the genes found in the sex determination QTL identified in V. emeryi (Miyakawa and Mikheyev 2015) (Supplementary Table S3). This approach also identified homology to the L. humile CSD locus, which was expected because a subset of V.emeryiCsdQTL2 is homologous to the CSD identified in L. humile (Pan et al. 2024). Homology to V.emeryiCsdQTL2 was scattered across one arm of O. biroi chromosome 4, including the 46kb CSD index peak, and homology to the L. humile CSD locus mapped within the O. biroi CSD index peak (Fig. 3A). The O. biroi CSD index peak contains four protein-coding genes that are also present in V.emeryiCsdQTL2, including HCF, COPA, an uncharacterized gene, and CRELD2. Notably, the candidate CSD region in O. biroi also includes a large stretch of sequence with no annotated genes downstream of an uncharacterized lncRNA. LncRNAs often retain their functions and synteny with surrounding genes despite divergence in sequence homology (Chodroff et al. 2010; Ulitsky et al. 2011; Quinn et al. 2016), raising the possibility that, although this lncRNA has limited sequence homology with L. humile ANTSR, it may have a homologous function (Pan et al. 2024). Regardless of the underlying molecular mechanism, our data suggest that this CSD locus is conserved between O. biroi, L. humile, and V. emeryi (Fig. 3B). This would mean that this locus is ancient, as these three species belong to different ant subfamilies that diverged roughly 112 million years ago (Borowiec et al. 2025).

The putative sex determination locus identified in O. biroi is conserved across formicoid ants.

(A) Karyoplot depicting the 14 chromosomes in the O. biroi genome, chromosome 4 from the L. humile genome (Pan et al. 2024), and one contig from linkage group 14 of the V. emeryi reference genome (Miyakawa and Mikheyev 2015). Note that the plots from the different species are not drawn to scale. Homology to protein-coding genes identified within V.emeryiCsdQTL2 is drawn in green. The location of the O. biroi CSD index peak is indicated. (B) A phylogeny of the ant subfamilies (adapted from Borowiec et al. 2019), with the presence of a sex determination locus homologous to the peak of our CSD index, or the absence of data shown. The yellow shaded background denotes the formicoid clade. (C) Karyoplot for O. biroi chromosome 4, depicting homology to V.emeryiCsdQTL2 and ancestral heterozygosity for six different clonal lines (A, B, C, D, I, and M). Grey histograms depict the number of ancestrally heterozygous SNPs in 100kb windows.

We performed our genetic mapping in clonal line A because this was the only clonal line in which we found diploid males (see below). However, different clonal lines of O. biroi vary in which portions of the genome are ancestrally homozygous (Oxley et al. 2014). Therefore, it was unclear whether heterozygosity at the mapped CSD locus is required for female development in all clonal lines of O. biroi. To investigate this, we inspected heterozygosity in diploid females from six different clonal lines by retrieving previously published genome sequences from clonal line B (Lacy et al. 2024) and sequencing genomes of diploid females from four additional clonal lines (from both the native and invasive ranges of O. biroi (Kronauer et al. 2012; Trible et al. 2020)). Females from all six clonal lines were heterozygous at the CSD index peak, consistent with its putative role as a CSD locus in all O. biroi (Fig. 3C).

The CSD index peak has high diversity in a non-coding region

Because CSD loci are expected to evolve under balancing selection, we expect many different alleles to exist at those loci. For example, as many as 19 csd alleles were found to segregate in honeybee populations (Hasselmann et al. 2008), and seven alleles were found to segregate in a population of L. humile, despite a recent genetic bottleneck (Pan et al. 2024). Looking for elevated genetic diversity is impossible by only studying genomes from a single clonal line (i.e., asexual descendants of a single diploid female), so we looked across diploid females from six different clonal lines. We calculated nucleotide diversity (l[) in 5kb sliding windows (step size = 1kb) across the genome and found a nucleotide diversity peak that fell within the CSD index peak (Fig. 4A, Supplementary Fig. S4), between positions 1,717,000 and 1,730,000 on chromosome 4 (Fig. 4B).

The CSD index peak is characterized by high genetic diversity in a non-coding region. (A &

B) Nucleotide diversity across the length of O. biroi chromosome 4 in 5kb windows (step size=1kb) (A) and across the vicinity of the CSD index peak in 100bp windows (step size=20bp) (B). (C) The number of differences per 100bp window between each alternate de novo assembled allele and the reference genome allele. (D) Annotated genes in the vicinity of the CSD index peak. Black boxes depict exons and thin lines depict introns. The lncRNA ANTSR is indicated in bold. Arrows indicate the names of genes in close proximity. The CSD index peak is shown as a green dotted line in A, and with green shading in B-D.

We hypothesized that this nucleotide diversity peak resulted from many alleles found among clonal lines of O. biroi at this locus. Identifying these alleles requires DNA sequences of individual haplotypes, whereas short-read genome sequences of diploid females only provide diploid genotypes. To investigate allelic diversity, we retrieved previously published haploid male genome sequences from clonal lines A and B (Lacy et al. 2024) and sequenced whole genomes of haploid males from additional clonal lines. We assembled their genomes de novo to identify the different haplotypes found in this region. In addition to the haplotype found in the reference genome, we identified six additional haplotypes (or “alleles”) in the five clonal lines for which we sequenced at least one haploid male genome (Fig. 4C, Table 2). These alleles differ substantially in their DNA sequences in the region with high nucleotide diversity (Fig. 4C).

To investigate which genetic elements might act to determine sex in a heterozygosity-dependent manner, we inspected all annotated genes in the CSD index peak (Fig. 4D). Within this region, none of the protein-coding genes contained substantial heterozygosity that affected amino acid identity (Supplementary File S3). To determine whether any genes or exons were missing from our annotation, we performed RNA sequencing of embryos and several other life stages (including female and male adults) using several different technologies (see Methods). Briefly, we used long-read RNA sequencing (PacBio IsoSeq) to extend gene models and used several different library preparation methods followed by standard short-read (Illumina) sequencing: standard poly-A tail library prep to capture mRNAs, ribosomal RNA depletion to capture lncRNAs that are not poly-adenylated, and small RNA sequencing to annotate miRNAs, endo-siRNAs, and putative piRNAs. Although this improved the O. biroi genome annotation by adding new genes and improving existing gene models (Supplementary Table S4, File S4), no new genes or exons were identified near the CSD index peak.

To look for sex-biased gene expression and splicing, we performed differential gene expression and differential transcript usage analysis between the transcriptomes of three male and three female early-stage (white) pupae. We used pupae for this experiment because this is the earliest life stage at which males and females can be readily distinguished. Primary sex determination signals act early during embryonic development (Sawanth et al. 2016) and thus need not be differentially expressed later in development. However, these genes can still be expressed as late as the pupal stage (Beye et al. 2003). Many genes were differentially expressed between males and females (24.6% of expressed genes) or had one or more differentially used exons, including four genes near the CSD index peak (LOC105285605, LOC105283850, LOC105283849, and LOC105283844) (Supplementary Tables S5 and S6; Fig. S5). However, none of these overlapped with the region of elevated genetic diversity. Very few reads from the male and female pupae samples aligned to the lncRNA, and although more reads aligned in RNA-seq libraries from eggs prepared using ribosomal RNA depletion, these data are insufficient for a rigorous assessment of sex-specific expression at this locus (Supplementary Fig. S6). Therefore, it remains unclear whether this lncRNA plays a role in sex determination in O. biroi.

The peak of nucleotide diversity minimally affects amino acid sequences and is primarily located in a non-coding genomic region (Fig. 4B-D). Thus, a non-coding genetic element within this region may instruct sex determination in O. biroi in a heterozygosity-dependent manner. This is similar to L. humile, where the mapped CSD locus had high variability in non-coding and non-exonic regions (Pan et al. 2024). RNA interference of the nearby lncRNA, ANTSR, led to male-specific splicing of tra, raising the possibility that heterozygosity or homozygosity in the non-exonic and non-coding L. humile CSD locus affects the expression level of ANTSR, which instructs sex-specific splicing of tra (Pan et al. 2024). Because the peak of nucleotide diversity within our CSD index peak minimally affects protein-coding sequences and the region is closely linked to a lncRNA that is putatively an ortholog of L. humile ANTSR, our data suggest that the mechanism of CSD in O. biroi may be conserved with L. humile.

O. biroi may have one or multiple sex determination loci

Our CSD index mapping results suggest that O. biroi has a single CSD locus rather than multiple loci. However, a curious observation raises the possibility that there may be other “ghost” CSD loci that went undetected in our mapping study. We sampled many males from clonal lines A and B, and a few males from four additional clonal lines. Despite this, all the diploid males we detected came from clonal line A. We, therefore, looked back at data from previously genotyped males and found that this pattern held across a previous study (Kronauer et al. 2012). In total, from clonal line A, 22 out of 114 males were diploid, whereas all 78 males genotyped from clonal line B were haploid (Table 3). This disparity is statistically significant (χ²=15.4, df=1, p<0.0001), suggesting that the two clonal lines differ in their propensity to produce diploid males.

One possible explanation for this phenomenon could be that O. biroi has multi-locus CSD (Fig. 1B), but that clonal line A is ancestrally homozygous for all but one of the loci, whereas clonal line B retains heterozygosity at two or more loci. Because multi-locus CSD requires homozygosity at all sex determination loci for diploids to develop as males, in this scenario, two or more losses of heterozygosity would be needed for diploid males to develop in clonal line B. In contrast, only one loss of heterozygosity would be required in clonal line A, which we used for our genetic mapping.

A second QTL region identified in V. emeryi (V.emeryiCsdQTL1) contains two closely linked tra homologs, similar to the closely linked honeybee tra homologs, csd and fem. This has led to the hypothesis that its function as a CSD locus is conserved with the csd-containing region of honeybees (Miyakawa and Mikheyev 2015). However, it remains to be demonstrated that either of these tra homologs acts as a primary CSD signal in V. emeryi. To test whether this locus might be involved in CSD in O. biroi, we searched for homology to V.emeryiCsdQTL1 in the O. biroi genome and identified several regions scattered across O. biroi chromosome 2 (Fig. 5A). However, we did not find a CSD index peak on chromosome 2, and no sex determination locus with homology to the V. emeryi QTL was found in L. humile (Pan et al. 2024). Thus, the function of V.emeryiCsdQTL1 as a primary sex determination signal might have evolved after this lineage separated from other ants (Fig. 5B). Alternatively, the sex determination activity of V.emeryiCsdQTL1 could be ancestral to these three ant species and independently lost in O. biroi (at least in clonal line A) and L. humile.

Whether a second, tra-containing CSD QTL from V. emeryi is conserved across ants remains ambiguous.

(A) Karyoplot depicting the 14 chromosomes in the O. biroi genome and one contig from linkage group 13 of the V. emeryi reference genome. Homology to protein-coding genes identified within V.emeryiCsdQTL1 is drawn in purple. The locations of the O. biroi CSD index peak and the homolog of transformer are indicated. (B) A phylogeny of the ant subfamilies (adapted from (Borowiec et al. 2019)), with the presence or absence of this second putative sex determination locus, or the absence of data shown. The yellow shaded background denotes the formicoid clade. (C) Karyoplot for O. biroi chromosome 2, depicting homology to V.emeryiCsdQTL1 and ancestral heterozygosity for each clonal line. Grey histograms depict the number of ancestrally heterozygous SNPs in 100kb windows. The region of homology that is ancestrally homozygous in clonal lines A and C is labeled.

Intriguingly, while clonal lines B, D, I, and M retain heterozygosity at all sites with homology to V.emeryiQTL1 (Fig. 5C), clonal lines A and C are homozygous in a region homologous to a portion of V.emeryiQTL1. We note, however, that heterozygosity in this region is limited, without elevated nucleotide diversity (Supplementary Fig. S4), and there are only two nonsynonymous heterozygous variants in this region, with clonal line B conspicuously lacking any nonsynonymous heterozygosity. Therefore, it seems unlikely that this region bears a CSD locus in O. biroi. We also note that all clonal lines are heterozygous in the region containing O. biroi’s tra ortholog (Fig. 5C), and that this gene is sex-specifically spliced (Supplementary Fig. S7), similar to tra homologs in other Holometabola (Hasselmann et al. 2008; Miyakawa and Miyakawa 2023). This contrasts with the tra-like gene in V. emeryi, which, like A. mellifera csd, is not sex-specifically spliced (Miyakawa and Miyakawa 2023).

Discussion

Genetically mapping sex determination loci in diverse taxa is key to understanding how sex determination evolves. Here, we investigated the mode of sex determination in the clonal raider ant, O. biroi. First, we mapped a sex determination locus that was heterozygous in all females, but homozygous for either allele in diploid males (Fig. 2). Next, we showed that most diploid males bear rare losses of heterozygosity that arise during thelytokous parthenogenesis (Fig. 2). We then showed that this locus is homologous to a sex determination locus found in two other ant species—V. emeryi and L. humile (Fig. 3), suggesting that this mechanism of sex determination may be conserved among formicoid ants. By sequencing whole genomes of haploid males and diploid females from other clonal lines of O. biroi, we showed that the mapped locus corresponds to a region with high genetic diversity, following the expectation that CSD loci should evolve under balancing selection (Fig. 4). This diversity peaks in a noncoding genomic region, and we found limited evidence for functional heterozygosity in nearby protein-coding sequences, suggesting that heterozygosity in this noncoding sequence may trigger female development. Finally, we demonstrated that O. biroi CSD in clonal line A does not map to the tra-containing sex determination locus previously identified in V. emeryi, but it remains unclear whether O. biroi ancestrally has single-locus or multi-locus CSD (Fig. 5).

The presence of diploid males in many ant taxa previously led to the conclusion that most ants employ CSD (van Wilgenburg et al. 2006). Along with L. humile (Pan et al. 2024) and V. emeryi (Miyakawa and Mikheyev 2015), O. biroi is now one of three ant species for which heterozygosity-dependent female development has been demonstrated at mapped CSD loci. Our mapped locus appears to be homologous to V.emeryiCsdQTL2 and to the locus identified in L. humile (Fig. 3). As was found in L. humile, this locus colocalizes with a peak of nucleotide diversity (Fig. 4), with seven alleles found among five clonal lines (Fig. 4C, Table 2), consistent with the expectation that CSD loci should harbor many alleles due to evolution under balancing selection. This implies that the role of this locus in CSD evolved before the divergence of the Dorylinae from the other formicoid ants roughly 112 million years ago (Borowiec et al. 2025). Although synteny analyses raised the possibility that this locus arose early within the evolution of the Aculeata (Pan et al. 2024), genetic mapping studies of additional representative species of different ant subfamilies (especially in the poneroid clade and the Leptanillinae) and across other Aculeata will be required to determine when the function of this locus in CSD originated and how it is distributed across taxa.

In L. humile, a lncRNA (ANTSR) closely linked to this locus plays a role in sex determination (Pan et al. 2024). Although we observed some expression of the putatively homologous lncRNA in O. biroi eggs of unknown sex, we were unable to assess sex-specific expression. The reason is that males in O. biroi are exceedingly rare and morphologically only identifiable at the pupal stage, well after primary sex determination signals are transduced to self-sustaining sex-specific splicing of downstream transcription factors during the early stages of embryonic development (Gempe and Beye 2011). Therefore, we likely missed differential gene expression relevant to sex determination that occurs before the pupal stage.

We did not find a peak of the CSD index at the O. biroi tra homolog (Fig. 5), indicating that, as in L. humile, the mode of CSD is not homologous to the system in A. mellifera mediated by the transformer homologs csd and fem. This was previously suspected for O. biroi due to the presence of only a single tra homolog (Oxley et al. 2014) rather than the two closely linked tra homologs found in other ants (Privman et al. 2013). In V. emeryi, however, one of the two Csd QTLs (V.emeryiCsdQTL1) spans the tra locus, raising the possibility that the molecular mechanism of CSD might, in part, be conserved between A. mellifera and V. emeryi (Miyakawa and Mikheyev 2015). It has not been demonstrated that either of V. emeryi’s tra homologs act as a primary sex determination signal. Our results raise the possibility that the role of V.emeryiCsdQTL1 as a primary sex determination signal evolved independently of A. mellifera csd and after V. emeryi diverged from L. humile. However, further investigations into the mechanisms of sex determination in V. emeryi and other ants are required to reconstruct the evolutionary history of V.emeryiCsdQTL1’s function in hymenopteran sex determination.

Curiously, diploid males were only produced in clonal line A, even though many males were collected from other clonal lines (Table 3). One possible explanation is that O. biroi has multi-locus CSD, and clonal line A is ancestrally homozygous for all but one of the loci. Clonal line A is ancestrally homozygous for a portion of the region homologous to the second V. emeryi QTL, although it retains heterozygosity in the region overlapping tra (Fig. 5C). However, this region lacks elevated genetic diversity and functional heterozygosity, making it a poor candidate for a second CSD locus. Further study would be required to determine whether O. biroi has a second CSD locus. However, there may be other explanations for diploid males being found exclusively in clonal line A. For example, clonal line A could be more likely to lose heterozygosity than other clonal lines, or diploid males in clonal line A could be more likely to develop and survive. Currently, we have no data to distinguish between these hypotheses.

Another peculiarity of our study is that five of the 16 sequenced diploid males retained heterozygosity at the mapped CSD locus. Stochastic gene expression or splicing fluctuations may have caused these individuals to develop as males despite having female genotypes. Certain allelic combinations of A. mellifera csd trigger female development with incomplete penetrance (Beye et al. 2013), and the allelic combination found in clonal line A might similarly be permissive to occasional male development. Alternatively, these diploid males could result from mutations in or losses of heterozygosity at different genetic loci. We note that it might be easier to find rare stochastically produced diploid males in laboratory colonies of O. biroi than in sexual species. This is because genetically encoded diploid males only result from loss of heterozygosity, which is rare in O. biroi (Oxley et al. 2014; Lacy et al. 2024). By contrast, in sexually reproducing populations, diploid males homozygous for sex determination loci will arise much more frequently due to matings between individuals that share an allele at the sex determination locus (this effect is exacerbated in invasive species like L. humile that have undergone recent genetic bottlenecks). Thus, if stochastically produced diploid males occur at very low baseline levels in ants, such males will rarely be detected in sexual species. In O. biroi, however, they could make up a substantial proportion of diploid males, simply because diploid males caused by homozygosity at the CSD locus are likewise extremely rare.

Finally, we note that heterozygosity-dependent female development, a fundamental tenet of CSD, has major implications for the evolution of breeding systems and reproductive modes. Due to the high fitness cost of producing sterile diploid males, species with CSD should obligatorily outbreed or have other mechanisms of maintaining heterozygosity (van Wilgenburg et al. 2006). In this study, most diploid males resulted from rare crossover-associated losses of heterozygosity that spanned the putative CSD locus (the shortest loss of heterozygosity at the mapped locus was 46kb and, therefore, likely resulted from crossover recombination rather than gene conversion). Intriguingly, in many taxa, asexual lineages reproduce via mechanisms that lead to complete homozygosity in a single generation (Suomalainen et al. 1987; Ma and Schwander 2017). However, among the social Hymenoptera (and other Vespoids), the known modes of asexual reproduction all maintain at least some heterozygosity (Rabeling and Kronauer 2012; Ma and Schwander 2017). For formicoid ants, bees, and other taxa with CSD, asexual reproduction with high rates of heterozygosity loss would incur a steep fitness penalty due to the production of sterile diploid males. This may help explain the unusual inheritance system recently described in O. biroi, where following crossover recombination, reciprocally recombined chromatids are faithfully co-inherited so that heterozygosity is rarely lost (Lacy et al. 2024).

How this ancient putative CSD locus triggers female development in a heterozygosity-dependent manner remains unknown. In L. humile, RNAi experiments suggested that the expression level of the lncRNA ANTSR affects sex-specific splicing of tra, even though heterozygosity in the exons of ANTSR is not required for this effect (Pan et al. 2024). Thus, the heterozygosity-dependent function of the putative CSD locus may depend on local regulatory interactions between homologous chromosomes, similar to a phenomenon known as transvection (Duncan 2002). To better understand how heterozygosity induces female development in ants, future studies will need to fine-map and functionally characterize this causal locus.

Materials and Methods

Software Versions

The versions of all software used for analyses are provided in Supplementary Table S7.

Identifying Male Ploidy

Clonal raider ant colonies are composed of female ants that reproduce asexually via a type of thelytokous parthenogenesis termed central fusion automixis (Oxley et al. 2014). Males are produced only sporadically (Oxley et al. 2014). Females of this species are diploid, whereas males can either be haploid (presumably arising via a failure of central fusion following female meiosis) or diploid (presumably arising due to rare losses of heterozygosity at CSD loci). Males are easily recognizable due to substantial sexual dimorphism (males have eyes and wings, whereas females are blind and flightless, and males also have different body shape and more darkly pigmented cuticles than females). We sampled them opportunistically whenever they were observed in colonies. We distinguished diploid males from haploid males based on heterozygosity at several unlinked genetic markers (Supplementary Table S1). For this, we disrupted one leg from each male using a Qiagen TissueLyser II and extracted DNA using Qiagen’s QIAmp DNA Micro Kit. We then PCR-amplified each marker and genotyped using Sanger sequencing (for most markers) or via gel electrophoresis either after restriction digestion (for two markers) or directly after amplification (for one marker at which the different alleles are different sizes). Details of the genotyping methods can be found in Supplementary Table S1.

Genome Sequencing and Preliminary Analysis

For each library, we disrupted an individual ant with a metal bead and Qiagen’s TissueLyser II and extracted DNA with Qiagen’s QIAmp DNA Micro Kit. We prepared Nextera Flex Illumina DNA sequencing libraries for 16 diploid males from clonal line A, six diploid females from different clonal lines, and seven haploid males from different clonal lines. We sequenced these libraries using an Illumina NovaSeq 6000. We also accessed libraries for individual genomes from previous studies. The information for all DNA sequencing libraries used in this study can be found in Supplementary Table S2.

We trimmed reads using Trimmomatic 0.36 (Bolger et al. 2014), aligned them to the O. biroi reference genome (McKenzie and Kronauer 2018) (Obir_v5.4, GenBank assembly accession: GCA_003672135.1) using bwa mem (Li 2013), and then sorted, deduplicated, and indexed using picard (http://broadinstitute.github.io/picard/). We called variants using GATK HaplotypeCaller (version 4.2) (Poplin et al. 2018). To exclude falsely collapsed regions in the reference genome, we filtered against sites found to be heterozygous in haploid males. We also screened out any variants for which putatively heterozygous samples had a proportionate minor allelic depth less than 0.25 and/or for which putatively homozygous samples had a minor allelic depth greater than zero. We performed custom filtering of variants and calculated statistics using pyvcf 0.6.8 (https://github.com/jamescasbon/PyVCF), and performed sliding window analyses using pybedtools 0.9.0 (Dale et al. 2011).

CSD Index Mapping

CSD loci are heterozygous in females but can be homozygous in diploid males. Thus, traditional association mapping cannot identify CSD loci because particular alleles do not trigger female development. Therefore, to map candidate CSD loci, we used a “CSD Index,” which equals zero if any female is homozygous, but if all females are heterozygous, it equals the proportion of diploid males that are homozygous. Before calculating this index, we had to identify ancestrally heterozygous sites. All individuals within an O. biroi clonal line are descended asexually from a single diploid female ancestor, and CSD loci would have been heterozygous in that common ancestor. Putatively ancestrally heterozygous variants are those for which at least one diploid individual is heterozygous, and all other diploid individuals are either heterozygous for the same two alleles, or homozygous for one of the two alleles for which other diploid(s) are heterozygous (Oxley et al. 2014; Lacy et al. 2024). After identifying putatively ancestrally heterozygous SNPs, we calculated the CSD index for each SNP.

To assess the likelihood that our observed CSD index values were due to chance (false positives), we randomly shuffled the identities of males and females to generate a null distribution of CSD index values for each SNP. In doing so, we maintained the underlying genetic structure by only shuffling sex labels within colonies. We computed p-values by calculating the proportion of permutations with values greater than or equal to the observed CSD index. To correct for multiple testing and control the false discovery rate across the genome-wide analysis, we applied the Benjamini-Hochberg False Discovery Rate correction and then recorded the mean CSD index value in sliding windows.

To directly evaluate the statistical association between homozygosity and sex, we applied Fisher’s Exact Test to assess whether SNP homozygosity frequencies differed significantly between males and females and used the Benjamini-Hochberg False Discovery Rate correction to adjust p-values.

Losses of Heterozygosity

We identified losses of heterozygosity as ancestrally heterozygous SNPs that had become homozygous in a given sample. To assess whether homozygosity for either allele at the mapped region permitted male development, we randomly chose a reference haploid male and assigned allelic identity based on whether a focal sample was homozygous for the same allele or the alternate allele to that possessed by the reference haploid male. To clearly illustrate losses of heterozygosity, we used the ordinal position of SNPs rather than the position of SNPs in the reference genome assembly.

Homozygosity in whole genome sequencing data can result from deletions, for which the affected region would have half the read depth of the genome-wide average, or copy-neutral losses of heterozygosity (which result rarely from meiotic recombination and central fusion parthenogenesis (Kronauer et al. 2012; Rabeling and Kronauer 2012; Oxley et al. 2014; Lacy et al. 2024)). To determine which of these processes caused homozygosity in diploid males, we identified runs of homozygosity using “bcftools roh” (Narasimhan et al. 2016), obtained read depths at all SNPs that passed filtering within these regions using “samtools depth –aa” (Danecek et al. 2021) and then normalized by dividing by the genome-wide median read depth. For samples without losses of heterozygosity on chromosome 4, we obtained normalized read depth at all SNPs that passed filtering on chromosome 4.

Homology Between CSD Loci Across Species

To determine whether our mapped CSD locus bore homology to those identified in other ant species, we used OrthoFinder (Emms and Kelly 2019) to identify orthologs of the genes present in the two CSD QTL in V. emeryi (Miyakawa and Mikheyev 2015). The locus identified in L. humile (Pan et al. 2024) is homologous to a subset of one of the V. emeryi QTL. In this process, we noticed that one of the genes found within our mapped CSD locus (cysteine-rich with EGF-like domain protein 2, or CRELD2) was also present in a second copy in a contig that was erroneously duplicated during the genome assembly process (Chr4:21571-39095). We masked this contig and realigned sequencing data for all subsequent analyses.

Balancing Selection

Because CSD loci are expected to evolve under negative frequency-dependent (balancing) selection, many different alleles should segregate in populations. To assess this, we used scikit-allel (Miles et al. 2019) to calculate nucleotide diversity across six different clonal lines and reported nucleotide diversity in sliding windows.

From inspecting DNA sequencing alignments to the reference genome, it became clear that the high genetic variation within the mapped CSD locus likely resulted from several highly differentiated haplotypes in this region. To investigate this further, we sequenced whole genomes of haploid males from clonal lines other than A and B. We assembled the genomes of each haploid male de novo using SPAdes (Bankevich et al. 2012; Prjibelski et al. 2020), found the contig bearing homology to the mapped CSD locus in the reference genome assembly using blastn (Camacho et al. 2009), manually scaffolded contigs as needed, and pairwise aligned each alternate allele to the reference genome using nucmer and mummer (Kurtz et al. 2004) to identify differences between alleles.

RNAseq for Annotation Improvements

To avoid missing any previously unannotated genes in the mapped CSD locus region, we performed RNA sequencing following various approaches. To start, we extracted total RNA from multiple samples, including separate pooled tissue from several different life stages (eggs, young larvae, fourth instar larvae, prepupae, pupae, and adults), different adult tissues (antennae, heads, thoraces, legs, abdomens), and individual male and female pupae using a TriZol RNA extraction method followed by precipitation in isopropanol at -20C overnight. We used the different life stage extractions for long read RNA-seq, rRNA depletion RNA-seq, and small RNA-seq. We used the different adult tissues for long read RNA-seq, and the individual male and female pupae for mRNA-seq. Long read RNA-seq was performed on a PacBio Sequel II, and all other sequencing was performed on an Illumina NovaSeq 6000. For rRNA depletion, we used Illumina’s TruSeq Stranded Total RNA RiboZero kit. For small RNA-seq, we prepared libraries using Perkin Elmer’s NEXTFLEX Small RNA-Seq Kit v3 for Illumina Platforms.

Annotation Improvements – Long Read Transcriptome Assembly

The GenBank reference genome (GCA_003672135) for O. biroi (Obir_v5.4) was retrieved from NCBI in FASTA format alongside the GenBank (GCA_003672135) and RefSeq (GCF_003672135.1) transcriptomes in GTF format. Additionally, a curated RefSeq annotation was retrieved from Zenodo (https://doi.org/10.5281/zenodo.10079884) in GTF format. RefSeq assemblies were chromosome name-mapped to GenBank chromosome names using the Bioconductor GenomeInfoDb package (Arora et al. 2024) and the chromosome name maps available in NCBI.

The PacBio toolset’s pbmm2 (https://github.com/PacificBiosciences/pbmm2), a PacBio wrapper for Minimap2 (Li 2018), was used for indexing and alignment of pooled full-length, non-concatemer long reads to the GenBank reference genome using the parameters “--preset ISOSEQ -j 8”. Following the generation of aligned long reads in BAM format, Stringtie2 (Kovaka et al. 2019) was used with parameters “-L” for long read transcriptome assembly, and the curated RefSeq annotation was supplied as a guide for assembly using the “-G” parameter. The resulting gene models were exported for further analysis and integration as a gzipped GTF file.

Annotation Improvements – Short Read Transcriptome Assembly

For Illumina RNA-seq data, FastQ quality control was performed using Rfastp (Wang and Carroll 2024). Indexing and alignment of short reads to the GenBank genome was performed using Hisat2 (Kim et al. 2019) with default parameters, and samtools was used to generate sorted and indexed BAM files for further analysis. Stringtie2 was then used to assemble transcriptomes using the default parameters. Gene models were then combined across replicates using Stringtie2’s merge function with parameters “-F 0.5 -T 2 -f 0.1” and were exported for further analysis and integration as a gzipped GTF file.

Annotation Improvements – Integrating Short and Long Read Transcriptomes with the RefSeq Assembly

Integration of long read and short read generated gene models with the established and curated RefSeq transcriptome was performed using custom scripts in R available from GitHub (https://github.com/RockefellerUniversity/Obiroi_GeneModels_2025) and described here.

First, conflicting gene models where single long read genes map to multiple RefSeq genes and vice versa were identified. As generated long read gene models are supported by multiple full-length, non-concatemer long read alignments, attention was focused on the 720 instances of multiple long read generated genes overlapping single RefSeq genes.

Single exon long read genes were removed as these are often found to be artifacts from PacBio sequencing. Following this, long read gene models that split RefSeq gene models but matched with GenBank gene models were accepted with the corresponding RefSeq models rejected.

Next, RefSeq gene models that conflicted within themselves were removed. This includes genes that were overlapping in their exons but marked as separate genes.

After tidying RefSeq gene models, long read gene models that split RefSeq gene models but matched with short read gene models were accepted, with the corresponding RefSeq models rejected. After this gene model polishing, only 132 gene models were left where multiple long read generated genes overlapped single RefSeq genes.

To potentially rescue internally primed transcripts caused by long stretches of genomic polyAs, long read transcripts were assessed for polyA nucleotide content downstream from their endpoints, and these distributions were visualized to identify cut-offs for putative internally primed transcripts. Transcripts with at least 70 percent as 10bp downstream of the last exon or with a polyA stretch > 4 in 10bp of the last exons were labeled as putative internally primed transcripts. For putative internally primed transcripts, which also split RefSeq genes, short read gene models were used to guide the rescued models when both short read and RefSeq agreed on a single gene model.

Finally, for the remaining long read gene models that split RefSeq gene models, gaps between long read gene models were assessed in the short read and RefSeq gene models. If agreement in exon and junction structure was found between short read and RefSeq gene models, the long read gene models were rejected. This left a final 27 genes where long read gene models split RefSeq gene models, and these were manually assessed in IGV alongside short read and long read alignments to reject or accept.

With a final set of long read gene models, all RefSeq and short read transcripts that fell within single genes were then added to the long read transcriptome to create an integrated transcriptome. Duplicate transcripts were then removed from the integrated transcriptome, and transcripts and genes were annotated for their sources (long read, short read, RefSeq) and any overlap with the established RefSeq, GenBank, and Ensembl gene models included. Following the creation of the integrated gene models, SQANTI3 (Pardo-Palacios et al. 2024) was used to provide a final corrected set of gene models as well as additional QC, as described below.

Assessment of Annotation Quality and Identifying Novel Genes, Transcripts and Exons

SQANTI3 was run with default parameters on the integrated transcriptome against the curated RefSeq and GenBank gene models to generate QC and descriptive statistics on novel and non-canonical splicing events. Gffcompare software (Pertea and Pertea 2020) was run against the curated RefSeq and GenBank gene models to provide statistics on novel exons, transcripts, and genes.

BUSCO (Simão et al. 2015) was run on the integrated, curated RefSeq and GenBank gene models with the parameters “-l insecta_odb10 -m tran” to provide measures of transcriptome completeness based on a set of core gene models. All these metrics were summarized into tables using custom R scripts available on GitHub (https://github.com/RockefellerUniversity/Obiroi_GeneModels_2025).

Annotation of CDS and Functional Annotation of Genes

Annotation of coding regions was performed using TransDecoder (Haas 2023) to add CDS information to our integrated gene models GTF. TransDecoder was run with default parameters except for providing the “--single_best_only” to the TransDecoder.Predict function to get only one prediction per transcript. CDS information was merged with SQANTI3’s GeneMark CDS annotation for a final set of CDSs. With this coding information, gene models were read into R, translated into protein sequences using the BioStrings package, and protein per transcript sequences exported as FASTA format for use in downstream tools. For functional annotation, both eggNOG (Huerta-Cepas et al. 2019) and InterProScan (Jones et al. 2014) were used using default parameters with our protein sequences FASTA to provide predictions for functional annotations of genes including their putative GO, KEGG, PFAM and Panther categories. Annotations were summarized to gene levels from transcript level annotation using R.

Annotation of lncRNAs

FEELnc (Wucher et al. 2017) was used to identify putative lncRNAs within our final integrated gene models. First, the FEELnc_filter.pl script was run with default parameters, and then the FEELnc_codpot.pl script was run with parameters “--mode=shuffle” to generate our putative set of lncRNAs for annotation of our integrated gene model GTF.

Annotation of miRNAs

Small RNA-seq reads were adaptor clipped for a “TGGAATTCTCGGGTGCCAAGG” adapter using CLIPflexR (Rozen-Gagnon et al. 2021) and aligned to the genome using miRDeep2’s mapper.pl script with parameters “-c -j -m -l 18” (Friedländer et al. 2012). Aligned data were then processed using miRDeep2’s miRDeep2.pl script with a FASTA of known Drosophila miRNAs retrieved from miRBase (Kozomara et al. 2019). The resulting annotated miRNA genomic locations were merged across samples and added into our integrated gene model GTF.

Annotation of piRNAs

For the identification of piRNA clusters, small read RNA-seq was processed using the pipeline workflow described within the Pilfer (Ray and Pandey 2018) toolkit to generate piRNA cluster BED files per sample. The sample piRNA clusters were then merged across samples to produce a final piRNA cluster set, which was then included in our integrated gene model GTF.

Annotation of tRNAs, snRNAs, snoRNAs and rRNAs

Additional annotations for tRNAs, snRNAs, snoRNAs and rRNAs were mapped from RefSeq annotation onto our integrated gene model GTF.

Sex-specific Gene Expression

To identify genes with sex-specific expression, we aligned Illumina mRNA-seq reads from three female and three male pupae using STAR (v2.5.0a) (Dobin et al. 2013). We counted and assigned reads to genes and transcripts using RSEM (v1.2.28) (Li and Dewey 2011). We analyzed differential gene expression using DESeq2 (Love et al. 2014), filtering out genes shorter than ten bases or with less than ten counts across all six samples. We hierarchically clustered the differentially expressed genes by Euclidian distance using the R package pheatmap (v1.0.12) (Kolde 2024). We used DEXSeq (v1.50.0) (Anders et al. 2012) to identify differential exon usage.

Data and Software Availability

All DNA and RNA sequencing data are publicly available at the National Center for Biotechnology Information Sequence Read Archive under accession number PRJNA1075055. All other data are available in the article and Supporting Information. All code is available on GitHub (genome annotation: https://github.com/RockefellerUniversity/Obiroi_GeneModels_2025; differential gene expression and exon usage: https://github.com/jina-leemon/CSD_biroi_male-female-RNA-seq; all other code and analyses: https://github.com/kipdlacy/SexDetermination_Obiroi).

eLife Assessment

This study provides valuable insights into the evolutionary conservation of sex determination mechanisms in ants by identifying a candidate sex-determining region in a parthenogenetic species. The strength of evidence is solid, using genomic analyses to identify differences in heterozygosity between females and diploid males, though the conclusions are limited by the lack of functional analysis.

Reviewer #1 (Public review):

This study investigates the sex determination mechanism in the clonal ant Ooceraea biroi, focusing on a candidate complementary sex determination (CSD) locus-one of the key mechanisms supporting haplodiploid sex determination in hymenopteran insects. Using whole genome sequencing, the authors analyze diploid females and the rarely occurring diploid males of O. biroi, identifying a 46 kb candidate region that is consistently heterozygous in females and predominantly homozygous in diploid males. This region shows elevated genetic diversity, as expected under balancing selection. The study also reports the presence of an lncRNA near this heterozygous region, which, though only distantly related in sequence, resembles the ANTSR lncRNA involved in female development in the Argentine ant, Linepithema humile (Pan et al. 2024). Together, these findings suggest a potentially conserved sex determination mechanism across ant species. However, while the analyses are well conducted and the paper is clearly written, the insights are largely incremental. The central conclusion - that the sex determination locus is conserved in ants - was already proposed and experimentally supported by Pan et al. (2024), who included O. biroi among the studied species and validated the locus's functional role in the Argentine ant. The present study thus largely reiterates existing findings without providing novel conceptual or experimental advances.

Other comments:

The mapping is based on a very small sample size: 19 females and 16 diploid males, and these all derive from a single clonal line. This implies a rather high probability for false-positive inference. In combination with the fact that only 11 out of the 16 genotyped males are actually homozygous at the candidate locus, I think a more careful interpretation regarding the role of the mapped region in sex determination would be appropriate. The main argument supporting the role of the candidate region in sex determination is based on the putative homology with the lncRNA involved in sex determination in the Argentine ant, but this argument was made in a previous study (as mentioned above).

In the abstract, it is stated that CSD loci have been mapped in honeybees and two ant species, but we know little about their evolutionary history. But CSD candidate loci were also mapped in a wasp with multi-locus CSD (study cited in the introduction). This wasp is also parthenogenetic via central fusion automixis and produces diploid males. This is a very similar situation to the present study and should be referenced and discussed accordingly, particularly since the authors make the interesting suggestion that their ant also has multi-locus CSD and neither the wasp nor the ant has tra homologs in the CSD candidate regions. Also, is there any homology to the CSD candidate regions in the wasp species and the studied ant?

The authors used different clonal lines of O. biroi to investigate whether heterozygosity at the mapped CSD locus is required for female development in all clonal lines of O. biroi (L187-196). However, given the described parthenogenesis mechanism in this species conserves heterozygosity, additional females that are heterozygous are not very informative here. Indeed, one would need diploid males in these other clonal lines as well (but such males have not yet been found) to make any inference regarding this locus in other lines.

Reviewer #2 (Public review):

The manuscript by Lacy et al. is well written, with a clear and compelling introduction that effectively conveys the significance of the study. The methods are appropriate and well-executed, and the results, both in the main text and supplementary materials, are presented in a clear and detailed manner. The authors interpret their findings with appropriate caution.

This work makes a valuable contribution to our understanding of the evolution of complementary sex determination (CSD) in ants. In particular, it provides important evidence for the ancient origin of a non-coding locus implicated in sex determination, and shows that, remarkably, this sex locus is conserved even in an ant species with a non-canonical reproductive system that typically does not produce males. I found this to be an excellent and well-rounded study, carefully analyzed and well contextualized.

That said, I do have a few minor comments, primarily concerning the discussion of the potential 'ghost' CSD locus. While the authors acknowledge (line 367) that they currently have no data to distinguish among the alternative hypotheses, I found the evidence for an additional CSD locus presented in the results (lines 261-302) somewhat limited and at times a bit difficult to follow. I wonder whether further clarification or supporting evidence could already be extracted from the existing data. Specifically:

(1) Line 268: I doubt the relevance of comparing the proportion of diploid males among all males between lines A and B to infer the presence of additional CSD loci. Since the mechanisms producing these two types of males differ, it might be more appropriate to compare the proportion of diploid males among all diploid offspring. This ratio has been used in previous studies on CSD in Hymenoptera to estimate the number of sex loci (see, for example, Cook 1993, de Boer et al. 2008, 2012, Ma et al. 2013, and Chen et al., 2021). The exact method might not be applicable to clonal raider ants, but I think comparing the percentage of diploid males among the total number of (diploid) offspring produced between the two lineages might be a better argument for a difference in CSD loci number.

(2) If line B indeed carries an additional CSD locus, one would expect that some females could be homozygous at the ANTSR locus but still viable, being heterozygous only at the other locus. Do the authors detect any females in line B that are homozygous at the ANTSR locus? If so, this would support the existence of an additional, functionally independent CSD locus.

(3) Line 281: The description of the two tra-containing CSD loci as "conserved" between Vollenhovia and the honey bee may be misleading. It suggests shared ancestry, whereas the honey bee csd gene is known to have arisen via a relatively recent gene duplication from fem/tra (10.1038/nature07052). It would be more accurate to refer to this similarity as a case of convergent evolution rather than conservation.

(4) Finally, since the authors successfully identified multiple alleles of the first CSD locus using previously sequenced haploid males, I wonder whether they also observed comparable allelic diversity at the candidate second CSD locus. This would provide useful supporting evidence for its functional relevance.

Overall, these are relatively minor points in the context of a strong manuscript, but I believe addressing them would improve the clarity and robustness of the authors' conclusions.

Reviewer #3 (Public review):

Summary:

The sex determination mechanism governed by the complementary sex determination (CSD) locus is one of the mechanisms that support the haplodiploid sex determination system evolved in hymenopteran insects. While many ant species are believed to possess a CSD locus, it has only been specifically identified in two species. The authors analyzed diploid females and the rarely occurring diploid males of the clonal ant Ooceraea biroi and identified a 46 kb CSD candidate region that is consistently heterozygous in females and predominantly homozygous in males. This region was found to be homologous to the CSD locus reported in distantly related ants. In the Argentine ant, Linepithema humile, the CSD locus overlaps with an lncRNA (ANTSR) that is essential for female development and is associated with the heterozygous region (Pan et al. 2024). Similarly, an lncRNA is encoded near the heterozygous region within the CSD candidate region of O. biroi. Although this lncRNA shares low sequence similarity with ANTSR, its potential functional involvement in sex determination is suggested. Based on these findings, the authors propose that the heterozygous region and the adjacent lncRNA in O. biroi may trigger female development via a mechanism similar to that of L. humile. They further suggest that the molecular mechanisms of sex determination involving the CSD locus in ants have been highly conserved for approximately 112 million years. This study is one of the few to identify a CSD candidate region in ants and is particularly noteworthy as the first to do so in a parthenogenetic species.

Strengths:

(1) The CSD candidate region was found to be homologous to the CSD locus reported in distantly related ant species, enhancing the significance of the findings.

(2) Identifying the CSD candidate region in a parthenogenetic species like O. biroi is a notable achievement and adds novelty to the research.

Weaknesses

(1) Functional validation of the lncRNA's role is lacking, and further investigation through knockout or knockdown experiments is necessary to confirm its involvement in sex determination.

(2) The claim that the lncRNA is essential for female development appears to reiterate findings already proposed by Pan et al. (2024), which may reduce the novelty of the study.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

This study investigates the sex determination mechanism in the clonal ant Ooceraea biroi, focusing on a candidate complementary sex determination (CSD) locus-one of the key mechanisms supporting haplodiploid sex determination in hymenopteran insects. Using whole genome sequencing, the authors analyze diploid females and the rarely occurring diploid males of O. biroi, identifying a 46 kb candidate region that is consistently heterozygous in females and predominantly homozygous in diploid males. This region shows elevated genetic diversity, as expected under balancing selection. The study also reports the presence of an lncRNA near this heterozygous region, which, though only distantly related in sequence, resembles the ANTSR lncRNA involved in female development in the Argentine ant, Linepithema humile (Pan et al. 2024). Together, these findings suggest a potentially conserved sex determination mechanism across ant species. However, while the analyses are well conducted and the paper is clearly written, the insights are largely incremental. The central conclusion - that the sex determination locus is conserved in ants - was already proposed and experimentally supported by Pan et al. (2024), who included O. biroi among the studied species and validated the locus's functional role in the Argentine ant. The present study thus largely reiterates existing findings without providing novel conceptual or experimental advances.

Although it is true that Pan et al., 2024 demonstrated (in Figure 4 of their paper) that the synteny of the region flanking ANTSR is conserved across aculeate Hymenoptera (including O. biroi), Reviewer 1’s claim that that paper provides experimental support for the hypothesis that the sex determination locus is conserved in ants is inaccurate. Pan et al., 2024 only performed experimental work in a single ant species (Linepithema humile) and merely compared reference genomes of multiple species to show synteny of the region, rather than functionally mapping or characterizing these regions.

Other comments:

The mapping is based on a very small sample size: 19 females and 16 diploid males, and these all derive from a single clonal line. This implies a rather high probability for false-positive inference. In combination with the fact that only 11 out of the 16 genotyped males are actually homozygous at the candidate locus, I think a more careful interpretation regarding the role of the mapped region in sex determination would be appropriate. The main argument supporting the role of the candidate region in sex determination is based on the putative homology with the lncRNA involved in sex determination in the Argentine ant, but this argument was made in a previous study (as mentioned above).

Our main argument supporting the role of the candidate region in sex determination is not based on putative homology with the lncRNA in L. humile. Instead, our main argument comes from our genetic mapping (in Fig. 2), and the elevated nucleotide diversity within the identified region (Fig. 4). Additionally, we highlight that multiple genes within our mapped region are homologous to those in mapped sex determining regions in both L. humile and Vollenhovia emeryi, possibly including the lncRNA.

In response to the Reviewer’s assertion that the mapping is based on a small sample size from a single clonal line, we want to highlight that we used all diploid males available to us. Although the primary shortcoming of a small sample size is to increase the probability of a false negative, small sample sizes can also produce false positives. We used two approaches to explore the statistical robustness of our conclusions. First, we generated a null distribution by randomly shuffling sex labels within colonies and calculating the probability of observing our CSD index values by chance (shown in Fig. 2). Second, we directly tested the association between homozygosity and sex using Fisher’s Exact Test (shown in Supplementary Fig. S2). In both cases, the association of the candidate locus with sex was statistically significant after multiple-testing correction using the Benjamini-Hochberg False Discovery Rate. These approaches are clearly described in the “CSD Index Mapping” section of the Methods.

We also note that, because complementary sex determination loci are expected to evolve under balancing selection, our finding that the mapped region exhibits a peak of nucleotide diversity lends orthogonal support to the notion that the mapped locus is indeed a complementary sex determination locus.

The fourth paragraph of the results and the sixth paragraph of the discussion are devoted to explaining the possible reasons why only 11/16 genotyped males are homozygous in the mapped region. The revised manuscript will include an additional sentence (in what will be lines 384-388) in this paragraph that includes the possible explanation that this locus is, in fact, a false positive, while also emphasizing that we find this possibility to be unlikely given our multiple lines of evidence.

In response to Reviewer 1’s suggestion that we carefully interpret the role of the mapped region in sex determination, we highlight our careful wording choices, nearly always referring to the mapped locus as a “candidate sex determination locus” in the title and throughout the manuscript. For consistency, the revised manuscript version will change the second results subheading from “The O. biroi CSD locus is homologous to another ant sex determination locus but not to honeybee csd” to “O. biroi’s candidate CSD locus is homologous to another ant sex determination locus but not to honeybee csd,” and will add the word “candidate” in what will be line 320 at the beginning of the Discussion, and will change “putative” to “candidate” in what will be line 426 at the end of the Discussion.

In the abstract, it is stated that CSD loci have been mapped in honeybees and two ant species, but we know little about their evolutionary history. But CSD candidate loci were also mapped in a wasp with multi-locus CSD (study cited in the introduction). This wasp is also parthenogenetic via central fusion automixis and produces diploid males. This is a very similar situation to the present study and should be referenced and discussed accordingly, particularly since the authors make the interesting suggestion that their ant also has multi-locus CSD and neither the wasp nor the ant has tra homologs in the CSD candidate regions. Also, is there any homology to the CSD candidate regions in the wasp species and the studied ant?

In response to Reviewer 1’s suggestion that we reference the (Matthey-Doret et al. 2019) study in the context of diploid males being produced via losses of heterozygosity during asexual reproduction, the revised manuscript will include (in what will be lines 123-126) the highlighted portion of the following sentence: “Therefore, if O. biroi uses CSD, diploid males might result from losses of heterozygosity at sex determination loci (Fig. 1C), similar to what is thought to occur in other asexual Hymenoptera that produce diploid males (Rabeling and Kronauer 2012; Matthey-Doret et al. 2019).”

We note, however, that in their 2019 study, Matthey-Doret et al. did not directly test the hypothesis that diploid males result from losses of heterozygosity at CSD loci during asexual reproduction, because the diploid males they used for their mapping study came from inbred crosses in a sexual population of that species.

We address this further below, but we want to emphasize that we do not intend to argue that O. biroi has multiple CSD loci. Instead, we suggest that additional, undetected CSD loci is one possible explanation for the absence of diploid males from any clonal line other than clonal line A. In response to Reviewer 1’s suggestion that we reference the (Matthey-Doret et al. 2019) study in the context of multilocus CSD, the revised manuscript version will include the following additional sentence in the fifth paragraph of the discussion (in what will be lines 372-374): “Multi-locus CSD has been suggested to limit the extent of diploid male production in asexual species under some circumstances (Vorburger 2013; Matthey-Doret et al. 2019).”

Regarding Reviewer 2’s question about homology between the putative CSD loci from the (Matthey-Doret et al. 2019) study and O. biroi, we note that there is no homology. The revised manuscript version will have an additional Supplementary Table (which will be the new Supplementary Table S3) that will report the results of this homology search. The revised manuscript will also include the following additional sentence in the Results, in what will be lines 172-174: “We found no homology between the genes within the O. biroi CSD index peak and any of the genes within the putative L. fabarum CSD loci (Supplementary Table S3).”

The authors used different clonal lines of O. biroi to investigate whether heterozygosity at the mapped CSD locus is required for female development in all clonal lines of O. biroi (L187-196). However, given the described parthenogenesis mechanism in this species conserves heterozygosity, additional females that are heterozygous are not very informative here. Indeed, one would need diploid males in these other clonal lines as well (but such males have not yet been found) to make any inference regarding this locus in other lines.

We agree that a full mapping study including diploid males from all clonal lines would be preferable, but as stated earlier in that same paragraph, we have only found diploid males from clonal line A. We stand behind our modest claim that “Females from all six clonal lines were heterozygous at the CSD index peak, consistent with its putative role as a CSD locus in all O. biroi.” In the revised manuscript version, this sentence (in what will be lines 199-201) will be changed slightly in response to a reviewer comment below: “All females from all six clonal lines (including 26 diploid females from clonal line B) were heterozygous at the CSD index peak, consistent with its putative role as a CSD locus in all O. biroi.”

Reviewer #2 (Public review):

The manuscript by Lacy et al. is well written, with a clear and compelling introduction that effectively conveys the significance of the study. The methods are appropriate and well-executed, and the results, both in the main text and supplementary materials, are presented in a clear and detailed manner. The authors interpret their findings with appropriate caution.

This work makes a valuable contribution to our understanding of the evolution of complementary sex determination (CSD) in ants. In particular, it provides important evidence for the ancient origin of a non-coding locus implicated in sex determination, and shows that, remarkably, this sex locus is conserved even in an ant species with a non-canonical reproductive system that typically does not produce males. I found this to be an excellent and well-rounded study, carefully analyzed and well contextualized.

That said, I do have a few minor comments, primarily concerning the discussion of the potential 'ghost' CSD locus. While the authors acknowledge (line 367) that they currently have no data to distinguish among the alternative hypotheses, I found the evidence for an additional CSD locus presented in the results (lines 261-302) somewhat limited and at times a bit difficult to follow. I wonder whether further clarification or supporting evidence could already be extracted from the existing data. Specifically:

We agree with Reviewer 2 that the evidence for a second CSD locus is limited. In fact, we do not intend to advocate for there being a second locus, but we suggest that a second CSD locus is one possible explanation for the absence of diploid males outside of clonal line A. In our initial version, we intentionally conveyed this ambiguity by titling this section “O. biroi may have one or multiple sex determination loci.” However, we now see that this leads to undue emphasis on the possibility of a second locus. In the revised manuscript, we will split this into two separate sections: “Diploid male production differs across O. biroi clonal lines” and “O. biroi lacks a tra-containing CSD locus.”

(1) Line 268: I doubt the relevance of comparing the proportion of diploid males among all males between lines A and B to infer the presence of additional CSD loci. Since the mechanisms producing these two types of males differ, it might be more appropriate to compare the proportion of diploid males among all diploid offspring. This ratio has been used in previous studies on CSD in Hymenoptera to estimate the number of sex loci (see, for example, Cook 1993, de Boer et al. 2008, 2012, Ma et al. 2013, and Chen et al., 2021). The exact method might not be applicable to clonal raider ants, but I think comparing the percentage of diploid males among the total number of (diploid) offspring produced between the two lineages might be a better argument for a difference in CSD loci number.

We want to re-emphasize here that we do not wish to advocate for there being two CSD loci in O. biroi. Rather, we want to explain that this is one possible explanation for the apparent absence of diploid males outside of clonal line A. We hope that the modifications to the manuscript described in the previous response help to clarify this.

Reviewer 2 is correct that comparing the number of diploid males to diploid females does not apply to clonal raider ants. This is because males are vanishingly rare among the vast numbers of females produced. We do not count how many females are produced in laboratory stock colonies, and males are sampled opportunistically. Therefore, we cannot report exact numbers. However, we will add the highlighted portion of the following sentence (in what will be lines 268-270) to the revised manuscript: “Despite the fact that we maintain more colonies of clonal line B than of clonal line A in the lab, all the diploid males we detected came from clonal line A.”

(2) If line B indeed carries an additional CSD locus, one would expect that some females could be homozygous at the ANTSR locus but still viable, being heterozygous only at the other locus. Do the authors detect any females in line B that are homozygous at the ANTSR locus? If so, this would support the existence of an additional, functionally independent CSD locus.

We thank the reviewer for this suggestion, and again we emphasize that we do not want to argue in favor of multiple CSD loci. We just want to introduce it as one possible explanation for the absence of diploid males outside of clonal line A.

The 26 sequenced diploid females from clonal line B are all heterozygous at the mapped locus, and the revised manuscript will clarify this in what will be lines 199-201. Previously, only six of those diploid females were included in Supplementary Table S2, and that will be modified accordingly.

(3) Line 281: The description of the two tra-containing CSD loci as "conserved" between Vollenhovia and the honey bee may be misleading. It suggests shared ancestry, whereas the honey bee csd gene is known to have arisen via a relatively recent gene duplication from fem/tra (10.1038/nature07052). It would be more accurate to refer to this similarity as a case of convergent evolution rather than conservation.

In the sentence that Reviewer 2 refers to, we are representing the assertion made in the (Miyakawa and Mikheyev 2015) paper in which, regarding their mapping of a candidate CSD locus that contains two linked tra homologs, they write in the abstract: “these data support the prediction that the same CSD mechanism has indeed been conserved for over 100 million years.” In that same paper, Miyakawa and Mikheyev write in the discussion section: “As ants and bees diverged more than 100 million years ago, sex determination in honey bees and V. emeryi is probably homologous and has been conserved for at least this long.”

As noted by Reviewer 2, this appears to conflict with a previously advanced hypothesis: that because fem and csd were found in Apis mellifera, Apis cerana, and Apis dorsata, but only fem was found in Mellipona compressipes, Bombus terrestris, and Nasonia vitripennis, that the csd gene evolved after the honeybee (Apis) lineage diverged from other bees (Hasselmann et al. 2008). However, it remains possible that the csd gene evolved after ants and bees diverged from N. vitripennis, but before the divergence of ants and bees, and then was subsequently lost in B. terrestris and M. compressipes. This view was previously put forward based on bioinformatic identification of putative orthologs of csd and fem in bumblebees and in ants [(Schmieder et al. 2012), see also (Privman et al. 2013)]. However, subsequent work disagreed and argued that the duplications of tra found in ants and in bumblebees represented convergent evolution rather than homology (Koch et al. 2014). Distinguishing between these possibilities will be aided by additional sex determination locus mapping studies and functional dissection of the underlying molecular mechanisms in diverse Aculeata.

Distinguishing between these competing hypotheses is beyond the scope of our paper, but the revised manuscript will include additional text to incorporate some of this nuance. We will include these modified lines below (in what will be lines 287-295), with the additions highlighted:

“A second QTL region identified in V. emeryi (V.emeryiCsdQTL1) contains two closely linked tra homologs, similar to the closely linked honeybee tra homologs, csd and fem (Miyakawa and Mikheyev 2015). This, along with the discovery of duplicated tra homologs that undergo concerted evolution in bumblebees and ants (Schmieder et al. 2012; Privman et al. 2013) has led to the hypothesis that the function of tra homologs as CSD loci is conserved with the csd-containing region of honeybees (Schmieder et al. 2012; Miyakawa and Mikheyev 2015). However, other work has suggested that tra duplications occurred independently in honeybees, bumblebees, and ants (Hasselmann et al. 2008; Koch et al. 2014), and it remains to be demonstrated that either of these tra homologs acts as a primary CSD signal in V. emeryi.”

(4) Finally, since the authors successfully identified multiple alleles of the first CSD locus using previously sequenced haploid males, I wonder whether they also observed comparable allelic diversity at the candidate second CSD locus. This would provide useful supporting evidence for its functional relevance.

As is already addressed in the final paragraph of the results and in Supplementary Fig. S4, there is no peak of nucleotide diversity in any of the regions homologous to V.emeryiQTL1, which is the tra-containing candidate sex determination locus (Miyakawa and Mikheyev 2015). In the revised manuscript, the relevant lines will be 307-310. We want to restate that we do not propose that there is a second candidate CSD locus in O. biroi, but we simply raise the possibility that multi-locus CSD *might* explain the absence of diploid males from clonal lines other than clonal line A (as one of several alternative possibilities).

Overall, these are relatively minor points in the context of a strong manuscript, but I believe addressing them would improve the clarity and robustness of the authors' conclusions.

Reviewer #3 (Public review):

Summary:

The sex determination mechanism governed by the complementary sex determination (CSD) locus is one of the mechanisms that support the haplodiploid sex determination system evolved in hymenopteran insects. While many ant species are believed to possess a CSD locus, it has only been specifically identified in two species. The authors analyzed diploid females and the rarely occurring diploid males of the clonal ant Ooceraea biroi and identified a 46 kb CSD candidate region that is consistently heterozygous in females and predominantly homozygous in males. This region was found to be homologous to the CSD locus reported in distantly related ants. In the Argentine ant, Linepithema humile, the CSD locus overlaps with an lncRNA (ANTSR) that is essential for female development and is associated with the heterozygous region (Pan et al. 2024). Similarly, an lncRNA is encoded near the heterozygous region within the CSD candidate region of O. biroi. Although this lncRNA shares low sequence similarity with ANTSR, its potential functional involvement in sex determination is suggested. Based on these findings, the authors propose that the heterozygous region and the adjacent lncRNA in O. biroi may trigger female development via a mechanism similar to that of L. humile. They further suggest that the molecular mechanisms of sex determination involving the CSD locus in ants have been highly conserved for approximately 112 million years. This study is one of the few to identify a CSD candidate region in ants and is particularly noteworthy as the first to do so in a parthenogenetic species.

Strengths:

(1) The CSD candidate region was found to be homologous to the CSD locus reported in distantly related ant species, enhancing the significance of the findings.

(2) Identifying the CSD candidate region in a parthenogenetic species like O. biroi is a notable achievement and adds novelty to the research.

Weaknesses

(1) Functional validation of the lncRNA's role is lacking, and further investigation through knockout or knockdown experiments is necessary to confirm its involvement in sex determination.

See response below.

(2) The claim that the lncRNA is essential for female development appears to reiterate findings already proposed by Pan et al. (2024), which may reduce the novelty of the study.

We do not claim that the lncRNA is essential for female development in O. biroi, but simply mention the possibility that, as in L. humile, it is somehow involved in sex determination. We do not have any functional evidence for this, so this is purely based on its genomic position immediately adjacent to our mapped candidate region. We agree with the reviewer that the study by Pan et al. (2024) decreases the novelty of our findings. Another way of looking at this is that our study supports and bolsters previous findings by partially replicating the results in a different species.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

L307-308 should state homozygous for either allele in THE MAJORITY of diploid males.

This will be fixed in the revised manuscript, in what will be line 321.

Reviewer #3 (Recommendations for the authors):

The association between heterozygosity in the CSD candidate region and female development in O. biroi, along with the high sequence homology of this region to CSD loci identified in two distantly related ant species, is not sufficient to fully address the evolution of the CSD locus and the mechanisms of sex determination.

Given that functional genetic tools, such as genome editing, have already been established in O. biroi, I strongly recommend that the authors investigate the role of the lncRNA through knockout or knockdown experiments and assess its impact on the sex-specific splicing pattern of the downstream tra gene.

Although knockout experiments of the lncRNA would be illuminating, the primary signal of complementary sex determination is heterozygosity. As is clearly stated in our manuscript and that of (Pan et al. 2024), it does not appear to be heterozygosity within the lncRNA that induces female development, but rather heterozygosity in non-transcribed regions linked to the lncRNA. Therefore, future mechanistic studies of sex determination in O. biroi, L. humile, and other ants should explore how homozygosity or heterozygosity of this region impacts the sex determination cascade, rather than focusing (exclusively) on the lncRNA.

With this in mind, we developed three sets of guide RNAs that cut only one allele within the mapped CSD locus, with the goal of producing deletions within the highly variable region within the mapped locus. This would lead to functional hemizygosity or homozygosity within this region, depending on how the cuts were repaired. We also developed several sets of PCR primers to assess the heterozygosity of the resultant animals. After injecting 1,162 eggs over several weeks and genotyping the hundreds of resultant animals with PCR, we confirmed that we could induce hemizygosity or homozygosity within this region, at least in ~1/20 of the injected embryos. Although it is possible to assess the sex-specificity of the splice isoform of tra as a proxy for sex determination phenotypes (as done by (Pan et al. 2024)), the ideal experiment would assess male phenotypic development at the pupal stage. Therefore, over several more weeks, we injected hundreds more eggs with these reagents and reared the injected embryos to the pupal stage. However, substantial mortality was observed, with only 12 injected eggs developing to the pupal stage. All of these were female, and none of them had been successfully mutated.

In conclusion, we agree with the reviewer that functional experiments would be useful, and we made extensive attempts to conduct such experiments. However, these experiments turned out to be extremely challenging with the currently available protocols. Ultimately, we therefore decided to abandon these attempts.

We opted not to include these experiments in the paper itself because we cannot meaningfully interpret their results. However, we are pleased that, in this response letter, we can include a brief description for readers interested in attempting similar experiments.

Since O. biroi reproduces parthenogenetically and most offspring develop into females, observing a shift from female- to male-specific splicing of tra upon early embryonic knockout of the lncRNA would provide much stronger evidence that this lncRNA is essential for female development. Without such functional validation, the authors' claim (lines 36-38) seems to reiterate findings already proposed by Pan et al. (2024) and, as such, lacks sufficient novelty.

We have responded to the issue of “lack of novelty” above. But again, the actual CSD locus in both O. biroi and L. humile appears to be distinct from (but genetically linked to) the lncRNA, and we have no experimental evidence that the putative lncRNA in O. biroi is involved in sex determination at all. Because of this, and given the experimental challenges described above, we do not currently intend to pursue functional studies of the lncRNA.

References

Hasselmann M, Gempe T, Schiøtt M, Nunes-Silva CG, Otte M, Beye M. 2008. Evidence for the evolutionary nascence of a novel sex determination pathway in honeybees. Nature 454:519–522.

Koch V, Nissen I, Schmitt BD, Beye M. 2014. Independent Evolutionary Origin of fem Paralogous Genes and Complementary Sex Determination in Hymenopteran Insects. PLOS ONE 9:e91883.

Matthey-Doret C, van der Kooi CJ, Jeffries DL, Bast J, Dennis AB, Vorburger C, Schwander T. 2019. Mapping of multiple complementary sex determination loci in a parasitoid wasp. Genome Biology and Evolution 11:2954–2962.

Miyakawa MO, Mikheyev AS. 2015. QTL mapping of sex determination loci supports an ancient pathway in ants and honey bees. PLOS Genetics 11:e1005656.

Pan Q, Darras H, Keller L. 2024. LncRNA gene ANTSR coordinates complementary sex determination in the Argentine ant. Science Advances 10:eadp1532.

Privman E, Wurm Y, Keller L. 2013. Duplication and concerted evolution in a master sex determiner under balancing selection. Proceedings of the Royal Society B: Biological Sciences 280:20122968.

Rabeling C, Kronauer DJC. 2012. Thelytokous parthenogenesis in eusocial Hymenoptera. Annual Review of Entomology 58:273–292.

Schmieder S, Colinet D, Poirié M. 2012. Tracing back the nascence of a new sex-determination pathway to the ancestor of bees and ants. Nature Communications 3:1–7.

Vorburger C. 2013. Thelytoky and Sex Determination in the Hymenoptera: Mutual Constraints. Sexual Development 8:50–58.

Version 2

Abstract

Abstract

Sex determination is a developmental switch that triggers sex-specific developmental programs. This switch is “flipped” by the expression of genes that promote male- or female-specific development. Many lineages have evolved sex chromosomes that act as primary signals for sex determination. However, haplodiploidy (males are haploid and females are diploid), which occurs in ca. 12% of animal species, is incompatible with sex chromosomes. Haplodiploid taxa must, therefore, rely on other strategies for sex determination. One mechanism, “complementary sex determination” (CSD), uses heterozygosity as a proxy for diploidy. In CSD, heterozygosity at a sex determination locus triggers female development, while hemizygosity or homozygosity permits male development. CSD loci have been mapped in honeybees and two ant species, but we know little about their evolutionary history. Here, we investigate sex determination in the clonal raider ant, Ooceraea biroi. We identified a 46kb candidate CSD locus at which all females are heterozygous, but most diploid males are homozygous for either allele. As expected for CSD loci, the candidate locus has more alleles than most other loci, resulting in a peak of nucleotide diversity. This peak negligibly affects the amino acid sequences of protein-coding genes, suggesting that heterozygosity of a non-coding genomic sequence triggers female development. This locus is distinct from the CSD locus in honeybees but homologous to a CSD locus mapped in two distantly related ant species, implying that this molecular mechanism has been conserved since a common ancestor that lived approximately 112 million years ago.

Introduction

Whether an animal develops as a male or a female is typically determined by a switch early in development. This switch, known as sex determination, is a multistep process beginning with primary sex determination signals triggering sex-specific splicing of downstream transcription factors that encode sexual identity (Williams and Carroll 2009). Although the downstream transcription factors are evolutionarily conserved, primary sex determination signals are evolutionarily labile and comprise a diverse range of environmental and genetic mechanisms (Bachtrog et al. 2014). Genetically encoded primary signals (sex-determining genes or chromosomes) are usually found in only one sex. Such systems do not work in haplodiploid taxa (an estimated 12% of animal species, including ants (Normark 2003)), where any allele on any homologous chromosome can be transmitted from diploid females to their haploid male sons, and all alleles experience selection to be viably transmitted from both sexes (Whiting 1935). Haplodiploids, therefore, require alternative sex determination mechanisms, but we know little about the molecular details.

Some parasitoid wasps use a mechanism involving maternal imprinting (Verhulst et al. 2010; Zou et al. 2020). However, in many Hymenoptera, inbred crosses produce diploid males, which led to the hypothesis that female development is triggered by heterozygosity as a proxy for diploidy (Whiting 1933; Whiting 1943). Under this “complementary sex determination” (CSD) hypothesis, different alleles at a given sex determination locus “complement” one another, meaning that heterozygosity triggers female development (Fig. 1A). By contrast, the presence of a single allele (either because the individual is haploid or because a diploid individual is homozygous at the sex determination locus) permits male development. Under CSD, diploid males produce diploid sperm and, therefore, are functionally sterile (van Wilgenburg et al. 2006). This has several evolutionary consequences (Cook and Crozier 1995); at the population level, there is selection against homozygosity, often meaning obligatory avoidance of inbreeding (Page 1980). At the gene level, there is negative frequency-dependent (balancing) selection on CSD loci, resulting in the co-occurrence of many alleles at similar frequencies within populations (Laidlaw et al. 1956).

Theoretical expectations for complementary sex determination (CSD).

(A) Cartoon depicting a mating between a diploid female with two different alleles at a sex determination locus and a haploid male bearing one of these two alleles. Half of the sexually produced diploid offspring are expected to develop as males. (B) Table illustrating multi-locus CSD in diploids, in which heterozygosity of at least one sex determination locus is required for female development. By contrast, only homozygosity at all loci results in diploid male development. (C) Cartoon depicting how CSD might work in asexual species such as the clonal raider ant. If diploid males arise from losses of heterozygosity at sex loci, then homozygosity for either of the two alleles should trigger male development. Offspring proportions reflect empirical observations in the clonal raider ant (Kronauer et al. 2012; Oxley et al. 2014).

Having been a favored hypothesis for more than half a century, CSD was demonstrated in the honeybee Apis mellifera in 2003 (Beye et al. 2003). In A. mellifera, biallelic heteromers of the protein encoded by the complementary sex determiner (csd) gene trigger female development, whereas monoallelic homomers permit male development (Beye et al. 2003; Beye et al. 2013; Otte et al. 2023; Seiler and Beye 2024). Although this primary sex determination signal differs from primary signals in other insects, it nonetheless converges upon the pathway that transduces sex determination signals in holometabolous insects (Bopp et al. 2014; Sawanth et al. 2016; Wexler et al. 2019). The Csd protein sex-specifically splices the closely linked splicing factor feminizer (fem) (the honeybee version of transformer (tra) in Drosophila and other insects), which in turn leads to the production of sex-specific isoforms of doublesex (dsx), a transcription factor that regulates sex-specific development (Hasselmann et al. 2008; Gempe et al. 2009; Verhulst et al. 2010; Zou et al. 2020). Intriguingly, csd is a paralog of fem. Following the prediction that CSD loci should evolve under balancing selection, many csd alleles are held at similar frequencies, and the locus bears signatures of adaptive evolution (Hasselmann and Beye 2004), with amino acid-level heterozygosity in certain protein domains required for female development (Otte et al. 2023).

CSD is assumed to occur in many Hymenoptera that occasionally produce diploid males (van Wilgenburg et al. 2006). Until recently, however, heterozygosity-dependent female development had not been demonstrated outside of A. mellifera, which remained the only species for which a CSD locus had been genetically mapped. A recent genetic mapping study in the wasp Lysiphlebus fabarum identified at least one and as many as four candidate sex determination loci (Matthey-Doret et al. 2019) (the multi-locus extension of the CSD hypothesis proposes that heterozygosity at any of multiple sex determination loci is sufficient to trigger female development (Crozier 1971) (Fig. 1B)). However, a requirement of heterozygosity at these loci for female development was not demonstrated. In the ant Vollenhovia emeryi, a mapping study found two sex determination QTL (Miyakawa and Mikheyev 2015), with follow-up work suggesting that these function as a multi-locus CSD system (Miyakawa and Miyakawa 2023). One of these QTL (V.emeryiCsdQTL1) contains two closely linked tra homologs, reminiscent of the locus containing csd and fem in A. mellifera. By contrast, the other QTL (V.emeryiCsdQTL2) only contains genes without annotated functions in sexual development. Although V. emeryi sex determination is likely integrated through tra and dsx, the molecular basis of their primary sex determination signals remains unclear (Miyakawa et al. 2018; Miyakawa and Miyakawa 2023).

Recently, a sex determination locus was identified in the Argentine ant, Linepithema humile (Pan et al. 2024). Consistent with CSD, all pairwise heterozygous combinations of the seven identified alleles were found in females, whereas diploid males were invariably homozygous for one of the seven alleles. Moreover, these alleles occurred at roughly equal frequencies within populations, suggesting evolution under balancing selection. The mapped locus is located within a noncoding genomic region tightly linked to ant noncoding transformer splicing regulator (ANTSR), a long noncoding RNA (lncRNA) with previously uncharacterized function. ANTSR knockdown in female-destined embryos leads to male-specific splicing of tra, suggesting that this lncRNA transduces the primary sex determination signal encoded by the heterozygosity (or hemi-or homozygosity) of the linked noncoding region (Pan et al. 2024). Because the synteny of this region (including the presence of a lncRNA putatively homologous to ANTSR) is conserved across the ants, bees, and vespoid wasps, it was suggested that this sex determination locus may be deeply conserved among the Aculeata (Pan et al. 2024). However, synteny is insufficient to demonstrate functional conservation, and genetic mapping of sex determination has yet to be performed in most of these taxa.

To explore the evolution of sex determination across ants, we investigated the topic in the clonal raider ant, Ooceraea biroi. Diploid males occur sporadically in this species, suggesting that, like other ants, it might employ CSD (Kronauer et al. 2012). O. biroi reproduces via a mode of parthenogenesis in which two haploid nuclei from a single meiosis fuse to produce diploid offspring (central fusion automixis) (Oxley et al. 2014). Although heterozygosity is maintained with high fidelity despite meiotic crossover recombination (Lacy et al. 2024), losses of heterozygosity occur occasionally, probably due to the inheritance of one recombined and one non-recombined version of a homologous chromosome (Kronauer et al. 2012; Oxley et al. 2014; Trible et al. 2023; Lacy et al. 2024). Therefore, if O. biroi uses CSD, diploid males might result from losses of heterozygosity at sex determination loci (Fig. 1C), similar to what is thought to occur in other asexual Hymenoptera that produce diploid males (Rabeling and Kronauer 2012; Matthey-Doret et al. 2019). Here, we use whole genome sequencing to map a sex determination locus, demonstrate that most diploid males carry a loss of heterozygosity at this locus, assess the homology of this locus with other sex determination loci, and explore whether balancing selection has shaped evolution at this locus.

Results

Identification of a candidate sex determination locus on chromosome 4

To map sex determination loci in O. biroi, we first identified males based on sexual dimorphism and distinguished diploid from haploid males based on heterozygosity at eight genetic markers (Supplementary Table S1, File S1, and Fig. S1). We then sequenced the genomes of 16 diploid males and compared them with 19 previously sequenced genomes of diploid females (Trible et al. 2023; Lacy et al. 2024) (descriptions and metadata for all genomes sequenced in this study are in Supplementary Table S2). CSD loci must be heterozygous to trigger female development but can be homozygous in diploid males. To identify such sites, we calculated a “CSD index” for each single nucleotide polymorphism (SNP), which equals zero if any female is homozygous and equals the proportion of diploid males that are homozygous if all females are heterozygous. We found a peak of the CSD index on chromosome 4, which deviated significantly from the null distribution of CSD index values (Fig. 2A). At this peak, homozygosity levels differed significantly between males and females (Supplementary Fig. S2).

Whole genome sequencing reveals a candidate sex determination locus on chromosome 4.

(A) Karyoplot depicting the mean CSD index p-value in 50kb windows with a 15kb sliding interval. The CSD index peak is shown as a green dotted line. The significance threshold (p=0.05) and FDR-corrected significance threshold (p=0.002) are indicated by gold and brown lines, respectively. The grey histogram shows the number of ancestrally heterozygous SNPs in 300kb windows. (B) Stacked bar plots for all diploids used for mapping, with one horizontal line for each ordered putatively ancestrally heterozygous SNP on chromosome 4. For each sample, SNPs that retain ancestral heterozygosity are drawn as grey lines, whereas SNPs that have lost ancestral heterozygosity are drawn as black lines.

To corroborate our results, we looked for evidence of segmental losses of heterozygosity along chromosome 4. Eleven out of 16 sequenced diploid males bore segmental losses of heterozygosity on chromosome 4 (Fig. 2B). The intersection of these LOH segments comprised a 46kb peak at which all females were heterozygous, and most (11 out of 16) diploid males were homozygous. Under the CSD hypothesis, homozygosity for any allele at CSD loci permits male development. Consistently, O. biroi diploid males were homozygous for either allele at the 46kb region on chromosome 4 (Table 1, Supplementary Fig. S3).

The stretches of homozygosity found in diploid males could have resulted from segmental deletions resulting from improperly repaired DNA damage or rare copy-neutral losses of heterozygosity resulting from thelytokous parthenogenesis (Kronauer et al. 2012; Oxley et al. 2014; Trible et al. 2023; Lacy et al. 2024). To distinguish between these two scenarios, we analyzed read depth and found that runs of homozygosity were not accompanied by changes in copy number (Supplementary Fig. S3). This implies that diploid males arise when rare losses of heterozygosity, which result from crossover recombination during meiosis, span the CSD locus.

Surprisingly, five of the 16 diploid males were not homozygous in this 46kb region. One possible explanation is that these individuals had mutations in other genes involved in the sex-determination pathway. We therefore inspected all unique mutations and losses of heterozygosity in these five diploid males, but did not find any in genes with annotated sex determining functions (Supplementary File S2). These individuals may carry mutations or losses of heterozygosity in genes or regulatory elements with unannotated sex determination function. Alternatively, male development may have been triggered by rare stochastic perturbations to gene expression or splicing. We address these possibilities further in the Discussion.

O. biroi’s candidate CSD locus is homologous to another ant sex determination locus but not to honeybee csd

To assess homology to CSD loci mapped in other species, we identified O. biroi orthologs of the genes found in other mapped sex determination loci. We found no homology between the genes within the O. biroi CSD index peak and any of the genes within the putative L. fabarum CSD loci (Supplementary Table S3). By identifying orthologs of genes within the two sex determination QTL identified in V. emeryi (Miyakawa and Mikheyev 2015) (Supplementary Table S4), we also identified homology to the L. humile CSD locus, because a subset of V.emeryiCsdQTL2 is homologous to the CSD identified in L. humile (Pan et al. 2024). Homology to V.emeryiCsdQTL2 was scattered across one arm of O. biroi chromosome 4, including the 46kb CSD index peak, and homology to the L. humile CSD locus mapped within the O. biroi CSD index peak (Fig. 3A). The O. biroi CSD index peak contains four protein-coding genes that are also present in V.emeryiCsdQTL2, including HCF, COPA, an uncharacterized gene, and CRELD2. Notably, the candidate CSD region in O. biroi also includes a large stretch of sequence with no annotated genes downstream of an uncharacterized lncRNA. LncRNAs often retain their functions and synteny with surrounding genes despite divergence in sequence homology (Chodroff et al. 2010; Ulitsky et al. 2011; Quinn et al. 2016), raising the possibility that, although this lncRNA has limited sequence homology with L. humile ANTSR, it may have a homologous function (Pan et al. 2024). Regardless of the underlying molecular mechanism, our data suggest that this CSD locus is conserved between O. biroi, L. humile, and V. emeryi (Fig. 3B). This would mean that this locus is ancient, as these three species belong to different ant subfamilies that diverged roughly 112 million years ago (Borowiec et al. 2025).

The putative sex determination locus identified in O. biroi is conserved across formicoid ants.

(A) Karyoplot depicting the 14 chromosomes in the O. biroi genome, chromosome 4 from the L. humile genome (Pan et al. 2024), and one contig from linkage group 14 of the V. emeryi reference genome (Miyakawa and Mikheyev 2015). Note that the plots from the different species are not drawn to scale. Homology to protein-coding genes identified within V.emeryiCsdQTL2 is drawn in green. The location of the O. biroi CSD index peak is indicated. (B) A phylogeny of the ant subfamilies (adapted from Borowiec et al. 2019), with the presence of a sex determination locus homologous to the peak of our CSD index, or the absence of data shown. The yellow shaded background denotes the formicoid clade. (C) Karyoplot for O. biroi chromosome 4, depicting homology to V.emeryiCsdQTL2 and ancestral heterozygosity for six different clonal lines (A, B, C, D, I, and M). Grey histograms depict the number of ancestrally heterozygous SNPs in 100kb windows.

We performed our genetic mapping in clonal line A because this was the only clonal line in which we found diploid males (see below). However, different clonal lines of O. biroi vary in which portions of the genome are ancestrally homozygous (Oxley et al. 2014). Therefore, it was unclear whether heterozygosity at the mapped CSD locus is required for female development in all clonal lines of O. biroi. To investigate this, we inspected heterozygosity in diploid females from six different clonal lines by retrieving previously published genome sequences from clonal line B (Lacy et al. 2024) and sequencing genomes of diploid females from four additional clonal lines (from both the native and invasive ranges of O. biroi (Kronauer et al. 2012; Trible et al. 2020)). All females from all six clonal lines (including 26 diploid females from clonal line B) were heterozygous at the CSD index peak, consistent with its putative role as a CSD locus in all O. biroi (Fig. 3C).

The CSD index peak has high diversity in a non-coding region

Because CSD loci are expected to evolve under balancing selection, we expect many different alleles to exist at those loci. For example, as many as 19 csd alleles were found to segregate in honeybee populations (Hasselmann et al. 2008), and seven alleles were found to segregate in a population of L. humile, despite a recent genetic bottleneck (Pan et al. 2024). Looking for elevated genetic diversity is impossible by only studying genomes from a single clonal line (i.e., asexual descendants of a single diploid female), so we looked across diploid females from six different clonal lines. We calculated nucleotide diversity (l[) in 5kb sliding windows (step size = 1kb) across the genome and found a nucleotide diversity peak that fell within the CSD index peak (Fig. 4A, Supplementary Fig. S4), between positions 1,717,000 and 1,730,000 on chromosome 4 (Fig. 4B).

The CSD index peak is characterized by high genetic diversity in a non-coding region. (A &

B) Nucleotide diversity across the length of O. biroi chromosome 4 in 5kb windows (step size=1kb) (A) and across the vicinity of the CSD index peak in 100bp windows (step size=20bp) (B). (C) The number of differences per 100bp window between each alternate de novo assembled allele and the reference genome allele. (D) Annotated genes in the vicinity of the CSD index peak. Black boxes depict exons and thin lines depict introns. The lncRNA ANTSR is indicated in bold. Arrows indicate the names of genes in close proximity. The CSD index peak is shown as a green dotted line in A, and with green shading in B-D.

We hypothesized that this nucleotide diversity peak resulted from many alleles found among clonal lines of O. biroi at this locus. Identifying these alleles requires DNA sequences of individual haplotypes, whereas short-read genome sequences of diploid females only provide diploid genotypes. To investigate allelic diversity, we retrieved previously published haploid male genome sequences from clonal lines A and B (Lacy et al. 2024) and sequenced whole genomes of haploid males from additional clonal lines. We assembled their genomes de novo to identify the different haplotypes found in this region. In addition to the haplotype found in the reference genome, we identified six additional haplotypes (or “alleles”) in the five clonal lines for which we sequenced at least one haploid male genome (Fig. 4C, Table 2). These alleles differ substantially in their DNA sequences in the region with high nucleotide diversity (Fig. 4C).

To investigate which genetic elements might act to determine sex in a heterozygosity-dependent manner, we inspected all annotated genes in the CSD index peak (Fig. 4D). Within this region, none of the protein-coding genes contained substantial heterozygosity that affected amino acid identity (Supplementary File S3). To determine whether any genes or exons were missing from our annotation, we performed RNA sequencing of embryos and several other life stages (including female and male adults) using several different technologies (see Methods). Briefly, we used long-read RNA sequencing (PacBio IsoSeq) to extend gene models and used several different library preparation methods followed by standard short-read (Illumina) sequencing: standard poly-A tail library prep to capture mRNAs, ribosomal RNA depletion to capture lncRNAs that are not poly-adenylated, and small RNA sequencing to annotate miRNAs, endo-siRNAs, and putative piRNAs. Although this improved the O. biroi genome annotation by adding new genes and improving existing gene models (Supplementary Table S5, File S4), no new genes or exons were identified near the CSD index peak.

To look for sex-biased gene expression and splicing, we performed differential gene expression and differential transcript usage analysis between the transcriptomes of three male and three female early-stage (white) pupae. We used pupae for this experiment because this is the earliest life stage at which males and females can be readily distinguished. Primary sex determination signals act early during embryonic development (Sawanth et al. 2016) and thus need not be differentially expressed later in development. However, these genes can still be expressed as late as the pupal stage (Beye et al. 2003). Many genes were differentially expressed between males and females (24.6% of expressed genes) or had one or more differentially used exons, including four genes near the CSD index peak (LOC105285605, LOC105283850, LOC105283849, and LOC105283844) (Supplementary Tables S5 and S6; Fig. S5). However, none of these overlapped with the region of elevated genetic diversity. Very few reads from the male and female pupae samples aligned to the lncRNA, and although more reads aligned in RNA-seq libraries from eggs prepared using ribosomal RNA depletion, these data are insufficient for a rigorous assessment of sex-specific expression at this locus (Supplementary Fig. S6). Therefore, it remains unclear whether this lncRNA plays a role in sex determination in O. biroi.

The peak of nucleotide diversity minimally affects amino acid sequences and is primarily located in a non-coding genomic region (Fig. 4B-D). Thus, a non-coding genetic element within this region may instruct sex determination in O. biroi in a heterozygosity-dependent manner. This is similar to L. humile, where the mapped CSD locus had high variability in non-coding and non-exonic regions (Pan et al. 2024). RNA interference of the nearby lncRNA, ANTSR, led to male-specific splicing of tra, raising the possibility that heterozygosity or homozygosity in the non-exonic and non-coding L. humile CSD locus affects the expression level of ANTSR, which instructs sex-specific splicing of tra (Pan et al. 2024). Because the peak of nucleotide diversity within our CSD index peak minimally affects protein-coding sequences and the region is closely linked to a lncRNA that is putatively an ortholog of L. humile ANTSR, our data suggest that the mechanism of CSD in O. biroi may be conserved with L. humile.

Diploid male production differs across O. biroi clonal lines.

Over the course of our study, we sampled many males from clonal lines A and B, and a few males from four additional clonal lines. Despite the fact that we maintain more colonies of clonal line B than of clonal line A in the lab, all the diploid males we detected came from clonal line A. We, therefore, looked back at data from previously genotyped males and found that this pattern held across a previous study (Kronauer et al. 2012). In total, from clonal line A, 22 out of 114 males were diploid, whereas all 78 males genotyped from clonal line B were haploid (Table 3). This disparity is statistically significant (χ²=15.4, df=1, p<0.0001), suggesting that the two clonal lines differ in their propensity to produce diploid males.

One possible explanation for this phenomenon could be that O. biroi has multi-locus CSD (Fig. 1B), but that clonal line A is ancestrally homozygous for all but one of the loci, whereas clonal line B retains heterozygosity at two or more loci. Because multi-locus CSD requires homozygosity at all sex determination loci for diploids to develop as males, in this scenario, two or more losses of heterozygosity would be needed for diploid males to develop in clonal line B. In contrast, only one loss of heterozygosity would be required in clonal line A, which we used for our genetic mapping. However, we identified no additional candidate loci in our mapping study, and other possibilities are equally likely, including that clonal lines A is more prone to loss of heterozygosity, or is more permissive to diploid male development than other clonal lines.

O. biroi lacks a tra-containing CSD locus

A second QTL region identified in V. emeryi (V.emeryiCsdQTL1) contains two closely linked tra homologs, similar to the closely linked honeybee tra homologs, csd and fem (Miyakawa and Mikheyev 2015). This, along with the discovery of duplicated tra homologs that undergo concerted evolution in bumblebees and ants (Schmieder et al. 2012; Privman et al. 2013) has led to the hypothesis that the function of tra homologs as CSD loci is conserved with the csd-containing region of honeybees (Schmieder et al. 2012; Miyakawa and Mikheyev 2015). However, other work has suggested that tra duplications occurred independently in honeybees, bumblebees, and ants (Hasselmann et al. 2008; Koch et al. 2014), and it remains to be demonstrated that either of these tra homologs acts as a primary CSD signal in V. emeryi. To test whether the tra homolog-containing QTL (V.emeryiCsdQTL1) might be involved in CSD in O. biroi, we searched for homology to this QTL in the O. biroi genome and identified several regions scattered across O. biroi chromosome 2 (Fig. 5A). However, we did not find a CSD index peak on chromosome 2, and no sex determination locus with homology to the V. emeryi QTL was found in L. humile (Pan et al. 2024). Thus, the function of V.emeryiCsdQTL1 as a primary sex determination signal might have evolved after this lineage separated from other ants (Fig. 5B). Alternatively, the sex determination activity of V.emeryiCsdQTL1 could be ancestral to these three ant species and independently lost in O. biroi (at least in clonal line A) and L. humile.

Whether a second, tra-containing CSD QTL from V. emeryi is conserved across ants remains ambiguous.

(A) Karyoplot depicting the 14 chromosomes in the O. biroi genome and one contig from linkage group 13 of the V. emeryi reference genome. Homology to protein-coding genes identified within V.emeryiCsdQTL1 is drawn in purple. The locations of the O. biroi CSD index peak and the homolog of transformer are indicated. (B) A phylogeny of the ant subfamilies (adapted from (Borowiec et al. 2019)), with the presence or absence of this second putative sex determination locus, or the absence of data shown. The yellow shaded background denotes the formicoid clade. (C) Karyoplot for O. biroi chromosome 2, depicting homology to V.emeryiCsdQTL1 and ancestral heterozygosity for each clonal line. Grey histograms depict the number of ancestrally heterozygous SNPs in 100kb windows. The region of homology that is ancestrally homozygous in clonal lines A and C is labeled.

Intriguingly, while clonal lines B, D, I, and M retain heterozygosity at all sites with homology to V.emeryiQTL1 (Fig. 5C), clonal lines A and C are homozygous in a region homologous to a portion of V.emeryiQTL1. We note, however, that heterozygosity in this region is limited, without elevated nucleotide diversity (Supplementary Fig. S4), and there are only two nonsynonymous heterozygous variants in this region, with clonal line B conspicuously lacking any nonsynonymous heterozygosity. Therefore, it seems unlikely that this region bears a CSD locus in O. biroi. We also note that all clonal lines are heterozygous in the region containing O. biroi’s tra ortholog (Fig. 5C), and that this gene is sex-specifically spliced (Supplementary Fig. S7), similar to tra homologs in other Holometabola (Hasselmann et al. 2008; Miyakawa and Miyakawa 2023). This contrasts with the tra-like gene in V. emeryi, which, like A. mellifera csd, is not sex-specifically spliced (Miyakawa and Miyakawa 2023).

Discussion

Genetically mapping sex determination loci in diverse taxa is key to understanding how sex determination evolves. Here, we investigated the mode of sex determination in the clonal raider ant, O. biroi. First, we mapped a candidate sex determination locus that was heterozygous in all females, but homozygous for either allele in most diploid males (Fig. 2). Next, we showed that most diploid males bear rare losses of heterozygosity that arise during thelytokous parthenogenesis (Fig. 2). We then showed that this locus is homologous to a sex determination locus found in two other ant species—V. emeryi and L. humile (Fig. 3), suggesting that this mechanism of sex determination may be conserved among formicoid ants. By sequencing whole genomes of haploid males and diploid females from other clonal lines of O. biroi, we showed that the mapped locus corresponds to a region with high genetic diversity, following the expectation that CSD loci should evolve under balancing selection (Fig. 4). This diversity peaks in a noncoding genomic region, and we found limited evidence for functional heterozygosity in nearby protein-coding sequences, suggesting that heterozygosity in this noncoding sequence may trigger female development. Finally, we demonstrated that O. biroi CSD in clonal line A does not map to the tra-containing sex determination locus previously identified in V. emeryi, but it remains unclear whether O. biroi ancestrally has single-locus or multi-locus CSD (Fig. 5).

The presence of diploid males in many ant taxa previously led to the conclusion that most ants employ CSD (van Wilgenburg et al. 2006). Along with L. humile (Pan et al. 2024) and V. emeryi (Miyakawa and Mikheyev 2015), O. biroi is now one of three ant species for which heterozygosity-dependent female development has been demonstrated at mapped CSD loci. Our mapped locus appears to be homologous to V.emeryiCsdQTL2 and to the locus identified in L. humile (Fig. 3). As was found in L. humile, this locus colocalizes with a peak of nucleotide diversity (Fig. 4), with seven alleles found among five clonal lines (Fig. 4C, Table 2), consistent with the expectation that CSD loci should harbor many alleles due to evolution under balancing selection. This implies that the role of this locus in CSD evolved before the divergence of the Dorylinae from the other formicoid ants roughly 112 million years ago (Borowiec et al. 2025).

Although synteny analyses raised the possibility that this locus arose early within the evolution of the Aculeata (Pan et al. 2024), genetic mapping studies of additional representative species of different ant subfamilies (especially in the poneroid clade and the Leptanillinae) and across other Aculeata will be required to determine when the function of this locus in CSD originated and how it is distributed across taxa.

In L. humile, a lncRNA (ANTSR) closely linked to this locus plays a role in sex determination (Pan et al. 2024). Although we observed some expression of the putatively homologous lncRNA in O. biroi eggs of unknown sex, we were unable to assess sex-specific expression. The reason is that males in O. biroi are exceedingly rare and morphologically only identifiable at the pupal stage, well after primary sex determination signals are transduced to self-sustaining sex-specific splicing of downstream transcription factors during the early stages of embryonic development (Gempe and Beye 2011). Therefore, we likely missed differential gene expression relevant to sex determination that occurs before the pupal stage.

We did not find a peak of the CSD index at the O. biroi tra homolog (Fig. 5), indicating that, as in L. humile, the mode of CSD is not homologous to the system in A. mellifera mediated by the transformer homologs csd and fem. This was previously suspected for O. biroi due to the presence of only a single tra homolog (Oxley et al. 2014) rather than the two closely linked tra homologs found in other ants (Privman et al. 2013). In V. emeryi, however, one of the two Csd QTLs (V.emeryiCsdQTL1) spans the tra locus, raising the possibility that the molecular mechanism of CSD might, in part, be conserved between A. mellifera and V. emeryi (Miyakawa and Mikheyev 2015). It has not been demonstrated that either of V. emeryi’s tra homologs act as a primary sex determination signal. Our results raise the possibility that the role of V.emeryiCsdQTL1 as a primary sex determination signal evolved independently of A. mellifera csd and after V. emeryi diverged from L. humile. However, further investigations into the mechanisms of sex determination in V. emeryi and other ants are required to reconstruct the evolutionary history of V.emeryiCsdQTL1’s function in hymenopteran sex determination.

Curiously, diploid males were only produced in clonal line A, even though many males were collected from other clonal lines (Table 3). One possible explanation is that O. biroi has multi-locus CSD, and clonal line A is ancestrally homozygous for all but one of the loci. Multi-locus CSD has been suggested to limit the extent of diploid male production in asexual species under some circumstances (Vorburger 2013; Matthey-Doret et al. 2019). Clonal line A is ancestrally homozygous for a portion of the region homologous to the second V. emeryi QTL, although it retains heterozygosity in the region overlapping tra (Fig. 5C). However, this region lacks elevated genetic diversity and functional heterozygosity, making it a poor candidate for a second, undetected, CSD locus. Further study would be required to determine whether O. biroi has a second CSD locus. However, there may be other explanations for diploid males being found exclusively in clonal line A. For example, clonal line A could be more likely to lose heterozygosity than other clonal lines, or diploid males in clonal line A could be more likely to develop and survive. Currently, we have no data to distinguish between these hypotheses.

Another peculiarity of our study is that five of the 16 sequenced diploid males retained heterozygosity at the mapped CSD locus. One possible explanation for this would be that our mapped locus is, in fact, a false positive. However, this seems unlikely given that 1) those losses of heterozygosity occurred independently in multiple genetic backgrounds within clonal line A (Supplementary Fig. S3), 2) a peak of nucleotide diversity co-occurs with the mapped locus (Fig. 4), and 3) this locus is homologous to CSD loci mapped in two other ant species (Fig. 3). Stochastic gene expression or splicing fluctuations may have caused these individuals to develop as males despite having female genotypes. Certain allelic combinations of A. mellifera csd trigger female development with incomplete penetrance (Beye et al. 2013), and the allelic combination found in clonal line A might similarly be permissive to occasional male development. Alternatively, these diploid males could result from mutations in or losses of heterozygosity at different genetic loci. We note that it might be easier to find rare stochastically produced diploid males in laboratory colonies of O. biroi than in sexual species. This is because genetically encoded diploid males only result from loss of heterozygosity, which is rare in O. biroi (Oxley et al. 2014; Lacy et al. 2024). By contrast, in sexually reproducing populations, diploid males homozygous for sex determination loci will arise much more frequently due to matings between individuals that share an allele at the sex determination locus (this effect is exacerbated in invasive species like L. humile that have undergone recent genetic bottlenecks). Thus, if stochastically produced diploid males occur at very low baseline levels in ants, such males will rarely be detected in sexual species. In O. biroi, however, they could make up a substantial proportion of diploid males, simply because diploid males caused by homozygosity at the CSD locus are likewise extremely rare.

Finally, we note that heterozygosity-dependent female development, a fundamental tenet of CSD, has major implications for the evolution of breeding systems and reproductive modes. Due to the high fitness cost of producing sterile diploid males, species with CSD should obligatorily outbreed or have other mechanisms of maintaining heterozygosity (van Wilgenburg et al. 2006). In this study, most diploid males resulted from rare crossover-associated losses of heterozygosity that spanned the putative CSD locus (the shortest loss of heterozygosity at the mapped locus was 46kb and, therefore, likely resulted from crossover recombination rather than gene conversion). Intriguingly, in many taxa, asexual lineages reproduce via mechanisms that lead to complete homozygosity in a single generation (Suomalainen et al. 1987; Ma and Schwander 2017). However, among the social Hymenoptera (and other Vespoids), the known modes of asexual reproduction all maintain at least some heterozygosity (Rabeling and Kronauer 2012; Ma and Schwander 2017). For formicoid ants, bees, and other taxa with CSD, asexual reproduction with high rates of heterozygosity loss would incur a steep fitness penalty due to the production of sterile diploid males. This may help explain the unusual inheritance system recently described in O. biroi, where following crossover recombination, reciprocally recombined chromatids are faithfully co-inherited so that heterozygosity is rarely lost (Lacy et al. 2024).

How this ancient putative CSD locus triggers female development in a heterozygosity-dependent manner remains unknown. In L. humile, RNAi experiments suggested that the expression level of the lncRNA ANTSR affects sex-specific splicing of tra, even though heterozygosity in the exons of ANTSR is not required for this effect (Pan et al. 2024). Thus, the heterozygosity-dependent function of the candidate CSD locus may depend on local regulatory interactions between homologous chromosomes, similar to a phenomenon known as transvection (Duncan 2002). To better understand how heterozygosity induces female development in ants, future studies will need to fine-map and functionally characterize this causal locus.

Materials and Methods

Software Versions

The versions of all software used for analyses are provided in Supplementary Table S8.

Identifying Male Ploidy

Clonal raider ant colonies are composed of female ants that reproduce asexually via a type of thelytokous parthenogenesis termed central fusion automixis (Oxley et al. 2014). Males are produced only sporadically (Oxley et al. 2014). Females of this species are diploid, whereas males can either be haploid (presumably arising via a failure of central fusion following female meiosis) or diploid (presumably arising due to rare losses of heterozygosity at CSD loci). Males are easily recognizable due to substantial sexual dimorphism (males have eyes and wings, whereas females are blind and flightless, and males also have different body shape and more darkly pigmented cuticles than females). We sampled them opportunistically whenever they were observed in colonies. We distinguished diploid males from haploid males based on heterozygosity at several unlinked genetic markers (Supplementary Table S1). For this, we disrupted one leg from each male using a Qiagen TissueLyser II and extracted DNA using Qiagen’s QIAmp DNA Micro Kit. We then PCR-amplified each marker and genotyped using Sanger sequencing (for most markers) or via gel electrophoresis either after restriction digestion (for two markers) or directly after amplification (for one marker at which the different alleles are different sizes). Details of the genotyping methods can be found in Supplementary Table S1.

Genome Sequencing and Preliminary Analysis

For each library, we disrupted an individual ant with a metal bead and Qiagen’s TissueLyser II and extracted DNA with Qiagen’s QIAmp DNA Micro Kit. We prepared Nextera Flex Illumina DNA sequencing libraries for 16 diploid males from clonal line A, six diploid females from different clonal lines, and seven haploid males from different clonal lines. We sequenced these libraries using an Illumina NovaSeq 6000. We also accessed libraries for individual genomes from previous studies. The information for all DNA sequencing libraries used in this study can be found in Supplementary Table S2.

We trimmed reads using Trimmomatic 0.36 (Bolger et al. 2014), aligned them to the O. biroi reference genome (McKenzie and Kronauer 2018) (Obir_v5.4, GenBank assembly accession: GCA_003672135.1) using bwa mem (Li 2013), and then sorted, deduplicated, and indexed using picard (http://broadinstitute.github.io/picard/). We called variants using GATK HaplotypeCaller (version 4.2) (Poplin et al. 2018). To exclude falsely collapsed regions in the reference genome, we filtered against sites found to be heterozygous in haploid males. We also screened out any variants for which putatively heterozygous samples had a proportionate minor allelic depth less than 0.25 and/or for which putatively homozygous samples had a minor allelic depth greater than zero. We performed custom filtering of variants and calculated statistics using pyvcf 0.6.8 (https://github.com/jamescasbon/PyVCF), and performed sliding window analyses using pybedtools 0.9.0 (Dale et al. 2011).

CSD Index Mapping

CSD loci are heterozygous in females but can be homozygous in diploid males. Thus, traditional association mapping cannot identify CSD loci because particular alleles do not trigger female development. Therefore, to map candidate CSD loci, we used a “CSD Index,” which equals zero if any female is homozygous, but if all females are heterozygous, it equals the proportion of diploid males that are homozygous. Before calculating this index, we had to identify ancestrally heterozygous sites. All individuals within an O. biroi clonal line are descended asexually from a single diploid female ancestor, and CSD loci would have been heterozygous in that common ancestor. Putatively ancestrally heterozygous variants are those for which at least one diploid individual is heterozygous, and all other diploid individuals are either heterozygous for the same two alleles, or homozygous for one of the two alleles for which other diploid(s) are heterozygous (Oxley et al. 2014; Lacy et al. 2024). After identifying putatively ancestrally heterozygous SNPs, we calculated the CSD index for each SNP.

To assess the likelihood that our observed CSD index values were due to chance (false positives), we randomly shuffled the identities of males and females to generate a null distribution of CSD index values for each SNP. In doing so, we maintained the underlying genetic structure by only shuffling sex labels within colonies. We computed p-values by calculating the proportion of permutations with values greater than or equal to the observed CSD index. To correct for multiple testing and control the false discovery rate across the genome-wide analysis, we applied the Benjamini-Hochberg False Discovery Rate correction and then recorded the mean CSD index value in sliding windows.

To directly evaluate the statistical association between homozygosity and sex, we applied Fisher’s Exact Test to assess whether SNP homozygosity frequencies differed significantly between males and females and used the Benjamini-Hochberg False Discovery Rate correction to adjust p-values.

Losses of Heterozygosity

We identified losses of heterozygosity as ancestrally heterozygous SNPs that had become homozygous in a given sample. To assess whether homozygosity for either allele at the mapped region permitted male development, we randomly chose a reference haploid male and assigned allelic identity based on whether a focal sample was homozygous for the same allele or the alternate allele to that possessed by the reference haploid male. To clearly illustrate losses of heterozygosity, we used the ordinal position of SNPs rather than the position of SNPs in the reference genome assembly.

Homozygosity in whole genome sequencing data can result from deletions, for which the affected region would have half the read depth of the genome-wide average, or copy-neutral losses of heterozygosity (which result rarely from meiotic recombination and central fusion parthenogenesis (Kronauer et al. 2012; Rabeling and Kronauer 2012; Oxley et al. 2014; Lacy et al. 2024)). To determine which of these processes caused homozygosity in diploid males, we identified runs of homozygosity using “bcftools roh” (Narasimhan et al. 2016), obtained read depths at all SNPs that passed filtering within these regions using “samtools depth –aa” (Danecek et al. 2021) and then normalized by dividing by the genome-wide median read depth. For samples without losses of heterozygosity on chromosome 4, we obtained normalized read depth at all SNPs that passed filtering on chromosome 4.

Homology Between CSD Loci Across Species

To determine whether our mapped CSD locus bore homology to those identified in other ant species, we used OrthoFinder (Emms and Kelly 2019) to identify orthologs of the genes present in the two CSD QTL in V. emeryi (Miyakawa and Mikheyev 2015). The locus identified in L. humile (Pan et al. 2024) is homologous to a subset of one of the V. emeryi QTL. In this process, we noticed that one of the genes found within our mapped CSD locus (cysteine-rich with EGF-like domain protein 2, or CRELD2) was also present in a second copy in a contig that was erroneously duplicated during the genome assembly process (Chr4:21571-39095). We masked this contig and realigned sequencing data for all subsequent analyses.

Balancing Selection

Because CSD loci are expected to evolve under negative frequency-dependent (balancing) selection, many different alleles should segregate in populations. To assess this, we used scikit-allel (Miles et al. 2019) to calculate nucleotide diversity across six different clonal lines and reported nucleotide diversity in sliding windows.

From inspecting DNA sequencing alignments to the reference genome, it became clear that the high genetic variation within the mapped CSD locus likely resulted from several highly differentiated haplotypes in this region. To investigate this further, we sequenced whole genomes of haploid males from clonal lines other than A and B. We assembled the genomes of each haploid male de novo using SPAdes (Bankevich et al. 2012; Prjibelski et al. 2020), found the contig bearing homology to the mapped CSD locus in the reference genome assembly using blastn (Camacho et al. 2009), manually scaffolded contigs as needed, and pairwise aligned each alternate allele to the reference genome using nucmer and mummer (Kurtz et al. 2004) to identify differences between alleles.

RNAseq for Annotation Improvements

To avoid missing any previously unannotated genes in the mapped CSD locus region, we performed RNA sequencing following various approaches. To start, we extracted total RNA from multiple samples, including separate pooled tissue from several different life stages (eggs, young larvae, fourth instar larvae, prepupae, pupae, and adults), different adult tissues (antennae, heads, thoraces, legs, abdomens), and individual male and female pupae using a TriZol RNA extraction method followed by precipitation in isopropanol at −20C overnight. We used the different life stage extractions for long read RNA-seq, rRNA depletion RNA-seq, and small RNA-seq. We used the different adult tissues for long read RNA-seq, and the individual male and female pupae for mRNA-seq. Long read RNA-seq was performed on a PacBio Sequel II, and all other sequencing was performed on an Illumina NovaSeq 6000. For rRNA depletion, we used Illumina’s TruSeq Stranded Total RNA RiboZero kit. For small RNA-seq, we prepared libraries using Perkin Elmer’s NEXTFLEX Small RNA-Seq Kit v3 for Illumina Platforms.

Annotation Improvements – Long Read Transcriptome Assembly

The GenBank reference genome (GCA_003672135) for O. biroi (Obir_v5.4) was retrieved from NCBI in FASTA format alongside the GenBank (GCA_003672135) and RefSeq (GCF_003672135.1) transcriptomes in GTF format. Additionally, a curated RefSeq annotation was retrieved from Zenodo (https://doi.org/10.5281/zenodo.10079884) in GTF format. RefSeq assemblies were chromosome name-mapped to GenBank chromosome names using the Bioconductor GenomeInfoDb package (Arora et al. 2024) and the chromosome name maps available in NCBI.

The PacBio toolset’s pbmm2 (https://github.com/PacificBiosciences/pbmm2), a PacBio wrapper for Minimap2 (Li 2018), was used for indexing and alignment of pooled full-length, non-concatemer long reads to the GenBank reference genome using the parameters “--preset ISOSEQ -j 8”. Following the generation of aligned long reads in BAM format, Stringtie2 (Kovaka et al. 2019) was used with parameters “-L” for long read transcriptome assembly, and the curated RefSeq annotation was supplied as a guide for assembly using the “-G” parameter. The resulting gene models were exported for further analysis and integration as a gzipped GTF file.

Annotation Improvements – Short Read Transcriptome Assembly

For Illumina RNA-seq data, FastQ quality control was performed using Rfastp (Wang and Carroll 2024). Indexing and alignment of short reads to the GenBank genome was performed using Hisat2 (Kim et al. 2019) with default parameters, and samtools was used to generate sorted and indexed BAM files for further analysis. Stringtie2 was then used to assemble transcriptomes using the default parameters. Gene models were then combined across replicates using Stringtie2’s merge function with parameters “-F 0.5 -T 2 -f 0.1” and were exported for further analysis and integration as a gzipped GTF file.

Annotation Improvements – Integrating Short and Long Read Transcriptomes with the RefSeq Assembly.

Integration of long read and short read generated gene models with the established and curated RefSeq transcriptome was performed using custom scripts in R available from GitHub (https://github.com/RockefellerUniversity/Obiroi_GeneModels_2025) and described here.

First, conflicting gene models where single long read genes map to multiple RefSeq genes and vice versa were identified. As generated long read gene models are supported by multiple full-length, non-concatemer long read alignments, attention was focused on the 720 instances of multiple long read generated genes overlapping single RefSeq genes.

Single exon long read genes were removed as these are often found to be artifacts from PacBio sequencing. Following this, long read gene models that split RefSeq gene models but matched with GenBank gene models were accepted with the corresponding RefSeq models rejected.

Next, RefSeq gene models that conflicted within themselves were removed. This includes genes that were overlapping in their exons but marked as separate genes.

After tidying RefSeq gene models, long read gene models that split RefSeq gene models but matched with short read gene models were accepted, with the corresponding RefSeq models rejected. After this gene model polishing, only 132 gene models were left where multiple long read generated genes overlapped single RefSeq genes.

To potentially rescue internally primed transcripts caused by long stretches of genomic polyAs, long read transcripts were assessed for polyA nucleotide content downstream from their endpoints, and these distributions were visualized to identify cut-offs for putative internally primed transcripts. Transcripts with at least 70 percent as 10bp downstream of the last exon or with a polyA stretch > 4 in 10bp of the last exons were labeled as putative internally primed transcripts. For putative internally primed transcripts, which also split RefSeq genes, short read gene models were used to guide the rescued models when both short read and RefSeq agreed on a single gene model.

Finally, for the remaining long read gene models that split RefSeq gene models, gaps between long read gene models were assessed in the short read and RefSeq gene models. If agreement in exon and junction structure was found between short read and RefSeq gene models, the long read gene models were rejected. This left a final 27 genes where long read gene models split RefSeq gene models, and these were manually assessed in IGV alongside short read and long read alignments to reject or accept.

With a final set of long read gene models, all RefSeq and short read transcripts that fell within single genes were then added to the long read transcriptome to create an integrated transcriptome. Duplicate transcripts were then removed from the integrated transcriptome, and transcripts and genes were annotated for their sources (long read, short read, RefSeq) and any overlap with the established RefSeq, GenBank, and Ensembl gene models included. Following the creation of the integrated gene models, SQANTI3 (Pardo-Palacios et al. 2024) was used to provide a final corrected set of gene models as well as additional QC, as described below.

Assessment of Annotation Quality and Identifying Novel Genes, Transcripts and Exons

SQANTI3 was run with default parameters on the integrated transcriptome against the curated RefSeq and GenBank gene models to generate QC and descriptive statistics on novel and non-canonical splicing events. Gffcompare software (Pertea and Pertea 2020) was run against the curated RefSeq and GenBank gene models to provide statistics on novel exons, transcripts, and genes.

BUSCO (Simão et al. 2015) was run on the integrated, curated RefSeq and GenBank gene models with the parameters “-l insecta_odb10 -m tran” to provide measures of transcriptome completeness based on a set of core gene models. All these metrics were summarized into tables using custom R scripts available on GitHub (https://github.com/RockefellerUniversity/Obiroi_GeneModels_2025).

Annotation of CDS and Functional Annotation of Genes

Annotation of coding regions was performed using TransDecoder (Haas 2023) to add CDS information to our integrated gene models GTF. TransDecoder was run with default parameters except for providing the “--single_best_only” to the TransDecoder.Predict function to get only one prediction per transcript. CDS information was merged with SQANTI3’s GeneMark CDS annotation for a final set of CDSs. With this coding information, gene models were read into R, translated into protein sequences using the BioStrings package, and protein per transcript sequences exported as FASTA format for use in downstream tools. For functional annotation, both eggNOG (Huerta-Cepas et al. 2019) and InterProScan (Jones et al. 2014) were used using default parameters with our protein sequences FASTA to provide predictions for functional annotations of genes including their putative GO, KEGG, PFAM and Panther categories.

Annotations were summarized to gene levels from transcript level annotation using R.

Annotation of lncRNAs

FEELnc (Wucher et al. 2017) was used to identify putative lncRNAs within our final integrated gene models. First, the FEELnc_filter.pl script was run with default parameters, and then the FEELnc_codpot.pl script was run with parameters “--mode=shuffle” to generate our putative set of lncRNAs for annotation of our integrated gene model GTF.

Annotation of miRNAs

Small RNA-seq reads were adaptor clipped for a “TGGAATTCTCGGGTGCCAAGG” adapter using CLIPflexR (Rozen-Gagnon et al. 2021) and aligned to the genome using miRDeep2’s mapper.pl script with parameters “-c -j -m −l 18” (Friedländer et al. 2012). Aligned data were then processed using miRDeep2’s miRDeep2.pl script with a FASTA of known Drosophila miRNAs retrieved from miRBase (Kozomara et al. 2019). The resulting annotated miRNA genomic locations were merged across samples and added into our integrated gene model GTF.

Annotation of piRNAs

For the identification of piRNA clusters, small read RNA-seq was processed using the pipeline workflow described within the Pilfer (Ray and Pandey 2018) toolkit to generate piRNA cluster BED files per sample. The sample piRNA clusters were then merged across samples to produce a final piRNA cluster set, which was then included in our integrated gene model GTF.

Annotation of tRNAs, snRNAs, snoRNAs and rRNAs

Additional annotations for tRNAs, snRNAs, snoRNAs and rRNAs were mapped from RefSeq annotation onto our integrated gene model GTF.

Sex-specific Gene Expression

To identify genes with sex-specific expression, we aligned Illumina mRNA-seq reads from three female and three male pupae using STAR (v2.5.0a) (Dobin et al. 2013). We counted and assigned reads to genes and transcripts using RSEM (v1.2.28) (Li and Dewey 2011). We analyzed differential gene expression using DESeq2 (Love et al. 2014), filtering out genes shorter than ten bases or with less than ten counts across all six samples. We hierarchically clustered the differentially expressed genes by Euclidian distance using the R package pheatmap (v1.0.12) (Kolde 2024). We used DEXSeq (v1.50.0) (Anders et al. 2012) to identify differential exon usage.

Data and Software Availability

All DNA and RNA sequencing data are publicly available at the National Center for Biotechnology Information Sequence Read Archive under accession number PRJNA1075055. All other data are available in the article and Supporting Information. All code is available on GitHub (genome annotation: https://github.com/RockefellerUniversity/Obiroi_GeneModels_2025; differential gene expression and exon usage: https://github.com/jina-leemon/CSD_biroi_male-female-RNA-seq; all other code and analyses: https://github.com/kipdlacy/SexDetermination_Obiroi).